\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}

Frontier language models have become the default backbone for agentic systems,
yet their dependence on external APIs rules them out for offline,
privacy-sensitive, or high-volume cost-constrained deployments. Small language models
(under 10B parameters) can run on consumer hardware, but without optimization, they
fail on the structured output and tool-calling tasks that agentic pipelines
require. This thesis investigates which optimization techniques are sufficient
to make a small language model reliably call tools, and at what accuracy cost
each technique comes.

An incremental ablation study is conducted on the Qwen~2.5 model family
(0.5B--7B), with \texttt{Qwen2.5-7B-Instruct} as the primary configuration,
evaluated on four BFCL v4 categories (\texttt{simple\_python},
\texttt{multiple}, \texttt{parallel}, \texttt{parallel\_multiple}) and
$\tau$-bench retail. The baseline, a minimal model-agnostic integration that
does not rely on the model's native tool-calling prompt template, scores 1.5\%,
approximately 77--96 percentage points below frontier models on the same
category (78.75\%--97.75\%); a control configuration that prompts the same model
through its native template with free generation reaches 96.00\%, inside the
frontier band, so the unoptimized gap measures integration quality, not model
capability. With constrained decoding alone, 72.75\% is reached without modifying model
weights or generating additional tokens; per-call latency under constrained
decoding is 0.878~s, compared with 6.995~s for unconstrained generation
(8$\times$ reduction). AWQ INT4 quantization retains 72.25\% while reducing
memory from 14.25~GiB to 5.20~GiB. Few-shot prompting, chain-of-thought
reasoning, and retrieval-augmented generation each reduce accuracy by 2.5,
6.75, and 24.5 percentage points relative to the constrained-decoding
configuration each of them extends. Format-aligned LoRA fine-tuning reaches
76.75\%, and 74.25\% after subsequent quantization at the same memory
footprint. Schema-enriched constrained decoding (CD+schema), which extends
the argument extraction prompt with parameter descriptions, enumeration
values, and default values while the FSM structural constraint is unchanged,
reaches 89.0\% ($+$16.5~pp over plain CD, McNemar $p < 0.00001$, $b{=}73$,
$c{=}7$) and places the 7B model inside the frontier range without
modifying model weights. On $\tau$-bench retail (115 tasks) and
\texttt{parallel} calls, accuracy falls to 4.3\% and 0\% respectively,
showing that single-call gains do not transfer to multi-turn or multi-call
settings.

The unoptimized performance gap is a format compliance gap, not a reasoning
gap: 394 of 400 failures in the unoptimized configuration are malformed outputs,
not wrong answers. This gap is resolved structurally by constrained decoding.
Production viability is assessed against two criteria. Frontier parity with lower-tier entries is reached by CD+schema:
89.0\% exceeds GPT-5-mini (78.75\%) and Gemini~3 Pro (79.58\%) while
falling 9~pp short of flagship models (97.75\%) (cross-harness
comparison against BFCL leaderboard scores, indicative only). The cascade criterion (a consumer GPU and a substantial
fraction of single-call traffic handled locally) is met by five
configurations, including CD+schema (89.0\%, 14.25~GiB) on a consumer NVIDIA
RTX~4090. Generic prompting interventions inside the pipeline cannot close the
residual failure rate after plain CD; CD+schema directly isolates schema
information loss as the dominant component, closing 16.5~pp of the residual
by restoring parameter descriptions, defaults, and enumerations that the
model-agnostic prompt discards. The LoRA training signal is retained in
attenuated form after AWQ INT4 quantization, although the residual gain over
the quantized base does not reach significance in a paired test; fine-tuned
capability thus remains obtainable on consumer hardware at INT4 precision.
