\chapter{Generative AI Disclosure}
\label{app:ai-disclosure}

In accordance with DTU's guidelines on the use of generative AI tools in
academic work, the following AI-assisted tools were used during the
preparation of this thesis.

\textbf{Claude Code (Anthropic)} was used as a coding assistant during
the implementation of the evaluation pipeline, HPC job scripts, and data
processing scripts. Generated code was reviewed, tested, and modified
before use. No generated code was submitted without verification against
expected outputs.

\textbf{Claude (Anthropic, claude.ai)} was used as a writing assistant
to produce draft text for thesis sections. All drafted text was reviewed,
revised, and rewritten in the author's own voice before inclusion. The
author is responsible for all factual claims, experimental results, and
interpretations presented in this thesis.

No AI tool was used to generate experimental data, benchmark results,
or citations. All numerical results reported in this thesis were produced
by running experiments on the DTU HPC cluster and collected by the author.

AI tools are not listed as authors or co-authors of this work, in
accordance with the Vancouver Convention as adopted by
DTU~\cite{ICMJE2020vancouver}.

\chapter{Hyperparameters and Training Details}
\label{app:hyperparameters}

Table~\ref{tab:lora-hyperparams} lists the hyperparameters used for
LoRA fine-tuning in all training configurations.

\begin{table}[h]
\centering
\caption{LoRA fine-tuning hyperparameters for all training configurations.}
\label{tab:lora-hyperparams}
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
Base model & \texttt{Qwen/Qwen2.5-7B-Instruct} \\
Dataset & \texttt{Salesforce/xlam-function-calling-60k} \\
Training examples & 54{,}000 \\
Validation examples & 6{,}000 \\
LoRA rank & 16 \\
LoRA alpha & 32 \\
Target modules & \texttt{q\_proj, v\_proj} \\
Dropout & 0.05 \\
Epochs & 2 \\
Batch size & 4 \\
Gradient accumulation steps & 4 \\
Learning rate & $2 \times 10^{-4}$ \\
LR scheduler & cosine \\
Dtype & bfloat16 \\
Optimizer & AdamW \\
Max sequence length & 2048 \\
\bottomrule
\end{tabular}
\end{table}

\chapter{Full Benchmark Results}
\label{app:benchmark-results}

Table~\ref{tab:full-results} reports the complete per-configuration results
on BFCL~v4 \texttt{simple\_python} (400 test cases).

\begin{table}[h]
\centering
\caption{Full BFCL v4 Simple Function AST accuracy results across all
configurations evaluated in this thesis.}
\label{tab:full-results}
\begin{tabular}{llrrl}
\toprule
Config & Description & Correct & Accuracy (↑) & Delta vs CD \\
\midrule
B       & No constrained decoding           & 6/400   & 1.50\%  & $-71.25$~pp \\
PE      & CD + few-shot (3 examples)        & 281/400 & 70.25\% & $-2.50$~pp  \\
CD      & Constrained decoding              & 291/400 & 72.75\% & baseline    \\
CD+Q    & CD + AWQ INT4                     & 289/400 & 72.25\% & $-0.50$~pp  \\
CD+Q+ITC & CD+Q + chain-of-thought         & 262/400 & 65.50\% & $-7.25$~pp  \\
CD+Q+RAG & CD+Q + FAISS top-5              & 191/400 & 47.75\% & $-25.00$~pp \\
FT-only & LoRA merged, no CD               & 55/400  & 13.75\% & $-59.00$~pp \\
CD+FT   & CD + LoRA merged (v1)            & 279/400 & 69.75\% & $-3.00$~pp  \\
FT-aligned-ng & Format-aligned LoRA, no CD   & 53/400  & 13.25\% & $-59.50$~pp \\
CD+FT-aligned & CD + format-aligned LoRA     & 307/400 & 76.75\% & $+4.00$~pp  \\
\bottomrule
\end{tabular}
\end{table}
