\chapter{Introduction} \label{ch:introduction}

\section{Motivation}

Large language models (LLMs) have become the dominant backbone for agentic systems. These systems reason over user requests, decompose tasks, and call external tools to produce structured outputs. Frontier models such as GPT and Claude perform strongly on these tasks, but they also introduce high inference costs, external API dependencies, added latency, and limited control over data privacy. In many deployment settings, these constraints are not secondary concerns; they determine whether the system can be used at all.

The dominant constraint depends on the setting. In on-device and edge applications, such as a phone assistant or an in-vehicle agent, there may be no reliable network path to an external API, and inference must fit within a fixed memory and power budget. In regulated domains such as healthcare, legal services, and finance, transmitting user data to a third-party API may be prohibited regardless of cost. In high-throughput backend automation, where a single task can issue many tool calls and request volume can reach millions per day, metered API calls dominate the operating budget. These settings share a common structure: many individual tool calls are mechanically simple. The model must emit one valid tool call with correct arguments given a tool schema and the current state, which is the schema- and API-constrained regime where SLMs are most competitive~\cite{sharma2025slmagents}. The practical question behind this thesis is whether a small local model can handle enough of these calls reliably, leaving only the hard cases for a frontier model.

Small language models (SLMs), commonly taken to span roughly 1 to 12 billion parameters~\cite{sharma2025slmagents}, offer a possible answer. Recent open-weight models such as Qwen 2.5~\cite{qwen2024qwen25}, Llama 3.2~\cite{metaai2024llama32}, and Phi-4~\cite{abdin2024phi} can run on consumer-grade hardware, support local deployment, and sharply reduce per-query costs. Out of the box, however, SLMs struggle with the capabilities that agentic systems require. They can produce malformed structured outputs, select incorrect tools, hallucinate function arguments, and fail at multi-step reasoning. These shortcomings have led to a widespread assumption that reliable agentic behavior requires large-scale models.

The failure mode is concrete, and the aggregate figures that follow
are measured on BFCL~v4 \texttt{simple\_python}. Consider a user query such as
\textit{``What is the sum of 265 and 345?''} directed at Qwen~2.5 7B-Instruct
without any optimization, where the available tool is an \texttt{add} function
that accepts two numeric arguments. The expected final output of the
evaluation pipeline is a single valid function call,
\texttt{\{"name": "add", "arguments": \{"a": 265, "b": 345\}\}}, assembled by
selecting the function name in the first stage and extracting its arguments
in the second. A logged generation from the argument-extraction stage
(temperature~0) instead produces the following, reproduced verbatim:
\begin{verbatim}
{"a": 265, "b": 345} The function `add` takes two numbers as
arguments and returns their sum as a string. [...] To compute the
sum, you would call the function like this:

```javascript
add(265, 345)
```

The result would be:

```json
"610"
```

So, the sum of 265 and 345 is 610.
\end{verbatim}
The model has identified the correct arguments---the leading
object \texttt{\{"a": 265, "b": 345\}} is itself well-formed---and the
correct tool, named in the explanatory text that follows, but that
trailing content makes the completion fail to parse as a single JSON object (a Python \texttt{json}
\texttt{Extra data} error at the first character past the object). On
BFCL~v4 \texttt{simple\_python}, 394 out of 400 responses from the
unoptimized 7B model fail in this way, yielding 1.5\% overall accuracy
(6/400). The problem is not that the model lacks the knowledge to answer;
it is that the model cannot reliably express that knowledge in the
structured format that agentic systems require.

This 1.5\% figure is the floor of a deliberately minimal,
model-agnostic integration, not the model's ceiling. It reflects
plain-text prompts, a strict single-object parse that discards the
leading valid object once trailing text follows it, and no use of the
model-specific tool-calling template that Qwen~2.5 provides. A more
lenient parser that accepts the leading valid object and ignores any
trailing text recovers many of these completions: the same 7B run then
scores 62.00\% (this lenient parser is the one used for the isolation
analysis in Chapter~\ref{ch:results}). A control
configuration that prompts the same model through its native
tool-calling template (B-template) reaches 96.00\% (384/400) on the
same test cases. The 1.5\% should therefore be read as the floor for a
generic integration rather than an intrinsic limit, and the
contribution of constrained decoding examined here is that it restores
format compliance structurally, with guarantees that hold for any model
and any schema. The full parser-sensitivity and B-template analysis is
deferred to Chapters~\ref{ch:results} and~\ref{ch:discussion}.

This thesis investigates one part of that assumption: whether reliable
single-call tool use requires a large model, or whether much of the observed
failure comes from insufficient structure around generation. It evaluates four
Qwen~2.5 Instruct sizes (0.5B, 1.5B, 3B, and 7B) on BFCL~v4
\texttt{simple\_python}, using a main ladder of optimization configurations and
additional isolation variants. The evaluation asks how much of the failure rate
is caused by output format, how much remains as semantic argument error, and
which interventions improve each part.

The main result is that constrained decoding removes format-compliance failures
at every model size and makes every output parseable under the required schema.
Against the strict generic floor, accuracy rises from 1.5--4.75\% in the
unoptimized setting to 51.50--72.75\%, depending on size, without modifying
model weights. At 7B, where the same unguided run is also evaluated under the
lenient first-object parser, the same-parser gain is smaller but still positive:
62.00\% to 72.75\% ($+$10.75~pp). The main contribution of constrained
decoding is therefore the model-agnostic structural guarantee, with an accuracy
gain whose apparent size depends on the parse contract. AWQ INT4 quantization
reduces memory by 63.5\% at the 7B size (from 14.25~GiB to 5.20~GiB) at a cost
of at most 0.5~pp above 1.5B, within run-to-run variance. Format-aligned LoRA
fine-tuning improves accuracy at every size; at 7B, CD+FT-aligned reaches
76.75\%, within single-run uncertainty of the weakest frontier entry on the
same category (GPT-5-mini, 78.75\%).

The strongest result is schema-enriched constrained decoding
(Config~CD+schema). This configuration adds parameter descriptions, enumeration
values, and defaults to the argument extraction prompt while keeping the FSM
mask unchanged. At 7B it reaches 89.00\% ($+$16.25~pp over plain CD, McNemar's
test $p < 0.00001$). This places the model in the lower frontier range on the
single-call, single-candidate BFCL~v4 Python Simple AST category, above
GPT-5-mini, without modifying model weights. The comparison is intentionally
narrow: it does not mean that the constrained pipeline is the best Qwen
integration, since the native-template control reaches 96.00\%, and it does
not imply general agent reliability. The failures that persist across
configurations and sizes are semantic rather than structural: wrong optional
parameter defaults, numeric precision mismatches, and string format conventions
that the model's learned priors do not match. This distinction, between failures
that constraints can fix and failures that require better semantic guidance or
training, organises the evaluation.

The practical deployment context motivating this investigation is the
cascade architecture: a two-tier system in which a locally deployed
small model handles routine requests, while a frontier model accessed
through an external API handles the remainder. This pattern is
economically motivated. Frontier model APIs are charged per token; a
locally deployed 7B model on a single consumer GPU incurs a fixed
hardware cost independent of query volume. If the small model handles
75\% of requests correctly and a router can identify the remaining cases,
75\% of per-query API calls can be avoided. The central challenge is extending
the fraction of requests that the small model can handle reliably and then
routing the residual semantic failures. This thesis evaluates that local SLM
tier in isolation for tool-calling tasks; it does not implement the cascade
router or escalation mechanism. BFCL accuracy is therefore used as an
upper-bound proxy for the fraction of single-call requests the SLM tier could
handle, providing a comparable basis for estimating cascade economics across
configurations rather than a measured deployment saving.

The optimization techniques are ordered by implementation cost. Methods
that require no modification to the model weights (constrained
decoding, prompt engineering, retrieval-augmented generation, and
post-training quantization) are evaluated first, as they can be
applied to any pre-trained checkpoint without additional compute.
LoRA fine-tuning is evaluated last, as it requires a training run on a
GPU cluster. This ordering reflects a practical decision ladder: a
deployment team can adopt the no-training configurations immediately
and defer the training investment until the marginal accuracy gain
justifies it.

\section{Problem Statement}

Agentic systems built on language models need two capabilities: they must reason correctly about user requests, and they must call the right tools with valid arguments. Reasoning covers task decomposition, the decision to call a tool or answer directly, and multi-step logic over intermediate results. Tool calling covers function selection, structured argument generation, and appropriate handling of the tool result. This thesis evaluates tool calling directly. Reasoning is measured only where it appears inside tool selection, argument construction, and one multi-step agentic benchmark; general-purpose reasoning benchmarks are outside the scope of the evaluation.

Without optimization, small language models fail at these tasks in several ways. They generate invalid JSON, select non-existent tools, produce arguments with incorrect types or values, and fail to follow execution plans. These failures are not merely inconvenient. In a production agent, they can prevent the system from calling tools at all.

Two categories of failure are distinguishable. Format failures occur
when the model produces syntactically invalid output: malformed JSON,
unclosed brackets, incorrect field names, or free-form text where a
structured object is required. These failures are independent of
whether the model understood the query; they reflect an inability to
express a correct answer in the required format. Semantic failures
occur when the output is structurally valid but contains incorrect
values: wrong function arguments, hallucinated parameters, or
mismatched types. These failures reflect gaps in the model's learned
associations between query context and argument conventions.

The distinction matters because the two failure types respond to
different interventions. Format failures can be eliminated at
inference time by constrained decoding, which masks invalid tokens at
each generation step without modifying the model. Semantic failures
cannot be addressed by structural constraints; they require either
training-based interventions that update the model's learned priors,
or retrieval-based methods that ground generation in retrieved
factual content. The central question is therefore not whether SLMs
fail, but which failure type dominates and which intervention
addresses it most effectively. The central finding of this thesis is
that constrained decoding converts the problem from format compliance
(fixable structurally at inference time) into semantic
argument-value accuracy, which only format-aligned training can
further improve.

Whether these failures stem from fundamental limitations in model
capacity or from insufficient constraints and guidance during
generation determines the appropriate intervention. If the latter,
a carefully chosen stack of optimization techniques should be
sufficient to make small models competitive with frontier models on
agentic benchmarks.

\section{Research Questions}

This thesis addresses three research questions:

\begin{description}
    \item[RQ1] \textit{What is the performance gap between unoptimized small language models (0.5B--7B parameters) and frontier models on tool-calling benchmarks?}
    \item[RQ2] \textit{What is the marginal contribution of each optimization technique (constrained decoding, schema enrichment, prompt engineering, chain-of-thought prompting, retrieval-augmented generation, quantization, and LoRA fine-tuning) when applied cumulatively to small models for tool calling?}
    \item[RQ3] \textit{Can an optimized small language model configuration provide an upper-bound local tier for cascade-style single-call tool calling on consumer hardware?}
\end{description}

\noindent RQ1 establishes the baseline gap that motivates the work. RQ2 is the core empirical contribution. It is answered through an ablation study with two parts: a cumulative ladder that adds one technique at a time, and a complementary isolation arm that evaluates inference-time techniques without constrained decoding. This separates each technique's standalone effect from the contribution of constrained decoding itself. RQ3 addresses the narrower deployment question of whether the optimized configuration can act as a local single-call tier in a cascade-style architecture.

For RQ3, the local-tier criterion is assessed against two explicit conditions.
The first is lower-tier frontier parity on the same benchmark category:
single-call accuracy within the range spanned by deployed frontier models on
BFCL~v4 Python Simple AST (78.75\%--97.75\%; the comparison set is given in
Chapter~\ref{ch:results}). The second is local deployability: a memory footprint
that fits a single consumer GPU (24~GiB), combined with single-call accuracy
high enough that the local tier could handle a substantial fraction of traffic
under an ideal semantic router ($p \geq 0.70$ is used as the reference point in
the cost model of Chapter~\ref{ch:discussion}). Both criteria are scoped to
single-call, schema-constrained tool calling. Neither covers general production
agents or multi-turn task completion, and the routing mechanism required to
escalate residual semantic failures is not implemented.

To answer these questions, the thesis evaluates an incremental pipeline of methods ordered by implementation cost. No-training methods are evaluated first: constrained decoding~\cite{willard2023efficient}, prompt engineering~\cite{brown2020language}, inference-time compute~\cite{wei2022chain}, retrieval-augmented generation~\cite{lewis2020retrieval}, and quantization~\cite{lin2024awq}. Training-based methods, represented by LoRA fine-tuning~\cite{hu2022lora}, are evaluated after the no-training methods. Each method is assessed both individually and in combination to measure its marginal contribution to tool-calling accuracy.

\section{Contributions}

The main contributions of this thesis are:

\begin{enumerate}
    \item A demonstration that constrained decoding alone gives a
    model-agnostic structural guarantee for BFCL tool-call generation, lifting
    the strict model-agnostic floor from 1.5--4.75\% to 51.50--72.75\% across
    the Qwen~2.5 size range (0.5B to 7B) without modifying model weights or
    relying on model-specific prompt templates. At 7B, where the same unguided
    run is also scored with the lenient first-object parser, constrained
    decoding improves accuracy from 62.00\% to 72.75\% (+10.75~pp) while
    making every output parseable.
    \item An incremental ablation study that isolates the marginal
    contribution of constrained decoding, prompt engineering,
    inference-time chain-of-thought, retrieval-augmented generation,
    quantization, LoRA fine-tuning, and schema-enriched prompting to
    BFCL \texttt{simple\_python} accuracy. Schema-enriched constrained
    decoding (CD+schema, 89.00\%) is the strongest no-training
    configuration and the first to enter the lower frontier range on the
    single-call, single-candidate BFCL~v4 Python Simple AST category
    (10.25~pp above GPT-5-mini, 78.75\%); CD+FT-aligned (76.75\%)
    approaches the same lower frontier tier.
    \item A scaling analysis across Qwen~2.5 0.5B, 1.5B, 3B, and 7B
    showing that the strict-parser recovery from constrained decoding scales
    with model size (48~pp at 0.5B to 71.25~pp at 7B), that the same-parser
    7B gain over the lenient unguided baseline is 10.75~pp, that quantization
    cost is negligible above 1.5B, that few-shot prompting reverses direction
    at 7B, and that format-aligned fine-tuning improves accuracy at every size.
    \item A characterization of the pipeline boundary and the schema change that
    crosses it: the two-stage single-call architecture achieves 70.5\% on the
    \texttt{multiple} category at 7B but 0\% on \texttt{parallel} by construction,
    and replacing the per-function extraction stage with a joint multi-call schema
    lifts \texttt{parallel} to 79\% at 7B, showing the limit was the output schema
    rather than model capability.
    \item An exploratory real-world validation on the GitHub MCP server (21 live
    tools, 50 natural-language prompts) showing that constrained decoding holds
    96.0\% tool selection at 7B when the model must choose from a full catalogue
    of semantically similar operations, that its benefit concentrates on the
    weakest viable model, and that forcing a choice over a large catalogue
    imposes a disambiguation cost on smaller models.
\end{enumerate}

\section{Thesis Structure}

The remainder of this thesis is organized as follows:

\textbf{Chapter~\ref{ch:background}} provides the theoretical background on language models, agentic systems, and the optimization techniques investigated in this work, including constrained decoding, prompt engineering, inference-time compute, retrieval-augmented generation, quantization, and LoRA fine-tuning.

\textbf{Chapter~\ref{ch:methodology}} describes the experimental design, the cumulative evaluation ladder, the candidate models, and the evaluation framework including benchmarks and metrics.

\textbf{Chapter~\ref{ch:results}} presents the experimental results for each optimization method, both individually and in combination, along with comparisons to frontier models.

\textbf{Chapter~\ref{ch:discussion}} discusses the implications of the results, analyzes the cost-benefit trade-offs of each method, and addresses limitations of the study.

\textbf{Chapter~\ref{ch:conclusion}} answers the research question, summarizes the findings, and outlines directions for future work.
