\chapter{Background} \label{ch:background}

Reliable tool calling with small language models draws on several areas of prior work. This chapter introduces the language-model architecture behind the evaluated systems, the design of agentic systems that invoke external tools, and the inference-time and training-time techniques used to control model output.

\section{Language Models}

\subsection{Transformer Architecture}

The Transformer architecture, introduced by Vaswani et al.~\cite{vaswani2017attention}, forms the foundation of modern language models. It replaces recurrent computation with self-attention, which allows each token in a sequence to attend to all other tokens in parallel. A Transformer block consists of a multi-head self-attention layer followed by a position-wise feed-forward network, with layer normalization and residual connections around each sub-layer.

Given an input sequence of token embeddings $X$, the self-attention mechanism computes queries $Q = XW^Q$, keys $K = XW^K$, and values $V = XW^V$ through learned linear projections. The attention output is computed as:
\begin{equation}
\text{Attention}(Q, K, V) = \text{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right)V
\end{equation}
where $d_k$ is the dimension of the key vectors. Multi-head attention runs this operation in parallel across multiple subspaces, allowing the model to capture different types of dependencies simultaneously.

For language modeling, the decoder-only variant is the most common
choice~\cite{brown2020language}, and has been shown to achieve the strongest
zero-shot generalization among architectures trained with unsupervised
pretraining~\cite{wang2022language}. It
applies a causal attention mask so that each token can only attend to
preceding tokens, enabling autoregressive generation where the model
predicts one token at a time conditioned on all previous tokens.

Efficient autoregressive generation is supported by the key-value
(KV) cache, which stores the key and value matrices computed for all
previous tokens so that they do not need to be recomputed at each
generation step. Without caching, generating a sequence of $T$ tokens
requires $O(T^2)$ attention operations; with caching, each new token
requires only $O(T)$ operations. The KV cache grows linearly with
sequence length and the number of attention heads, and its memory
footprint becomes a significant constraint at long context lengths
or large batch sizes.

Grouped-query attention (GQA) reduces KV cache memory by sharing a
single set of key and value heads across multiple query heads. In
standard multi-head attention, each of the $h$ query heads has a
corresponding key and value head, resulting in $2h$ matrices stored
in the KV cache per layer. With GQA, $h$ query heads are divided into
$g$ groups, each sharing one key and one value head, reducing the
cache to $2g$ matrices per layer where $g \ll h$. GQA is used in the
Qwen~2.5 model family~\cite{qwen2024qwen25} and is a key reason why
the 7B model fits within consumer GPU memory budgets during inference.

\subsection{Scaling Laws}

Scaling laws describe how model performance changes with the number of parameters, the amount of training data, and the available compute budget. Kaplan et al.~\cite{kaplan2020scaling} showed that language model loss follows a power-law relationship with these factors, and that performance improves predictably as they increase.

Hoffmann et al.~\cite{hoffmann2022training} refined these findings with the Chinchilla scaling laws, showing that many large models were under-trained relative to their size. For a fixed compute budget, performance improves when model size and training data are balanced rather than when parameter count is maximized alone. This matters for small language models because a well-trained small model can outperform a larger but under-trained one.

More recently, the focus has shifted toward inference-time scaling~\cite{snell2024scaling}, where additional computation at generation time (through techniques such as chain-of-thought prompting or search) can improve performance without increasing model size.

\subsection{Small vs. Large Language Models}

Large language models (LLMs) such as GPT-4~\cite{openai2023gpt4} and the Claude model family operate at hundreds of billions of parameters and perform strongly across a wide range of tasks, including complex reasoning and tool use. Their size also makes them expensive to run, typically requiring multi-GPU clusters or cloud API access for inference.

Small language models (SLMs), commonly taken to span roughly 1 to 12 billion parameters~\cite{sharma2025slmagents}, can run on consumer hardware such as a single GPU with 24GB of VRAM. Recent open-weight SLMs, including Qwen 2.5~\cite{qwen2024qwen25}, Llama 3.2~\cite{metaai2024llama32}, and Phi-4~\cite{abdin2024phi}, have narrowed the gap with larger models on standard benchmarks. This improvement comes from better training data, improved tokenizers, and architectural refinements such as grouped-query attention (GQA).

Despite these improvements, SLMs still lag behind frontier models on tasks that require multi-step reasoning, tool selection from large schemas, and reliable structured output generation. This gap motivates the optimization techniques explored in this thesis.

\section{Agentic Systems}

An agentic system is a software architecture in which a language model acts as the central reasoning engine. It decides which actions to take, invokes external tools, and synthesizes results to fulfill user requests. Unlike simple question-answering, agentic behavior involves planning, acting, and adapting based on intermediate results.

\subsection{Tool Calling}

Tool calling is the ability of a language model to invoke external functions or APIs as part of its response. The model receives a set of available tool schemas, including names, descriptions, and parameter types. It must then select the appropriate tool for a request, generate valid arguments matching the schema, and incorporate the tool's response into its reasoning.

Tool calling is the primary capability evaluated in this thesis. It requires both language understanding (to map user intent to the correct tool) and structured generation (to produce syntactically and semantically valid function calls).

\subsection{Structured Output Generation}

Agentic systems typically require the model to produce outputs in a specific format, most commonly JSON. The output must satisfy three conditions: it must be syntactically valid, conform to a predefined schema, and contain semantically correct values for the given context.

Without explicit constraints, language models frequently produce malformed outputs: missing closing brackets, incorrect field names, wrong data types, or extra fields not present in the schema. These failures are particularly common in small models and represent a key reliability bottleneck for agentic deployments.

\subsection{The ReAct Pattern}

ReAct (Reasoning and Acting)~\cite{yao2023react} is a prompting framework that interleaves reasoning traces with action steps. Rather than generating a final answer directly, the model follows a loop of Thought, Action, and Observation. The Thought step reasons about what to do next, the Action step calls a tool or performs a step, and the Observation step processes the result.

This pattern improves multi-step task completion by making the model's reasoning explicit and allowing it to adapt its plan based on intermediate results. ReAct is particularly relevant for agentic systems because it provides a structured way to combine reasoning and tool use within a single generation process.

Within the ReAct loop, the Action step is where the model must emit a structured tool call. For small models, this is also where generation often fails. The reasoning trace may be sound, but the emitted call may be malformed JSON or may name a function that does not exist.

This thesis isolates that step. Constrained decoding (Section~\ref{sec:bg-constrained-decoding}) enforces structural validity at the Action step, so that a small model's reasoning is not discarded because of a formatting error. Chapter~\ref{ch:results} evaluates whether this shifts the bottleneck from structural failure to residual semantic failure.

\subsection{Cascade Architectures}

A cascade architecture routes inference requests across two or
more models of increasing capability and cost. Routine requests
are handled by a smaller, faster model deployed locally; requests
that exceed its reliable operating range are escalated to a larger
model, typically accessed through an external API. The routing
decision may be based on query complexity, output confidence, or
post-hoc structural validation of the smaller model's response.

For tool-calling tasks, structural validity is a necessary but not
sufficient condition for accepting a small model's response. A
response that is syntactically valid JSON and names an existing
function may still contain incorrect argument values. The residual
semantic failure rate after structural constraints are enforced
therefore determines the escalation boundary: queries whose failure
mode is argument value errors cannot be filtered by structural
checks alone.

This thesis is framed around extending the small model operating
region within a cascade; the economic motivation and deployment
context are described in Chapter~\ref{ch:introduction}.

\section{Constrained Decoding} \label{sec:bg-constrained-decoding}

Constrained decoding modifies generation so that the model can only produce outputs that conform to a predefined structure. Instead of relying on the model to learn output formats implicitly, constrained decoding enforces them at inference time by masking invalid tokens at each generation step.

\subsection{Formal Grammars for Generation}

The key idea is to define the valid output space with a formal grammar, typically a context-free grammar (CFG) or a JSON schema. At each decoding step, the grammar determines which tokens are valid continuations given the tokens generated so far. Invalid tokens receive probability zero, or negative infinity in log-space, so only grammatically valid sequences can be produced.

For tool calling, this means the model is guaranteed to produce valid JSON that conforms to the tool's argument schema. Field names, types, and structure are enforced by construction rather than learned behavior. Willard and Louf~\cite{willard2023efficient} formalized this approach using finite-state machines compiled from regular expressions and parsers for context-free grammars.

A context-free grammar specifies the valid output space as a set of production rules. A minimal grammar for a JSON object carrying a single integer field illustrates the idea:

\begin{verbatim}
    object -> "{" '"a":' integer "}"
    integer -> digit | digit integer
    digit  -> "0" | "1" | ... | "9"
\end{verbatim}

\noindent After the opening brace, the grammar admits only the literal \texttt{"a":}, so any token whose text does not extend that literal is masked. After the field marker, it admits a digit, and this continues until the closing brace. The per-step overhead is low because the partial output is not re-parsed at every step. Instead, the grammar is compiled once, before generation, into a finite-state machine. The model's tokenizer vocabulary is then aligned against that machine, so each state stores the token IDs that are valid continuations from it~\cite{willard2023efficient}.

At generation time, the constraint is therefore a lookup of the current state's precomputed token mask. The cost does not grow with the size of the grammar; XGrammar~\cite{dong2024xgrammar} further caches these masks to reduce the remaining per-step cost. This indexing step is what libraries such as Outlines~\cite{outlines2024} perform, and it is why constrained decoding can enforce a non-trivial grammar with little added latency per token.

\subsection{Guided Decoding Techniques}

Several libraries implement constrained decoding for language models. Outlines~\cite{outlines2024} uses finite-state machines compiled from regular expressions or JSON schemas to guide generation. LMQL~\cite{beurer2023lmql} combines constrained decoding with a query language for expressing complex output constraints declaratively. XGrammar~\cite{dong2024xgrammar} accelerates context-free grammar execution with a token-mask cache, reducing the per-step overhead of grammar-constrained generation.

Format constraints are not free in general. Tam et al.~\cite{tam2024letmespeak} report that strict format restrictions can degrade reasoning performance on tasks where free-form derivation helps. Chapter~\ref{ch:results} examines this trade-off empirically for tool calling.

These techniques are particularly impactful for small models because they eliminate an entire class of errors (malformed outputs) without requiring any additional training. Constrained decoding is the base layer of this thesis, upon which the other techniques are stacked.

\section{Prompt Engineering and Inference-Time Compute}

\subsection{Few-Shot Prompting}

Few-shot prompting provides the model with examples of correct input-output pairs directly in the prompt~\cite{brown2020language}. For tool calling, this means including examples of user requests paired with the expected function calls. The model uses these examples as templates to guide its own generation.

Few-shot prompting is a zero-cost technique (no training required) that can improve task performance on format-learning and pattern-matching tasks. Its effectiveness depends on whether the bottleneck is output format compliance or semantic convention matching; for the latter, in-context examples may not generalise across function domains.

\subsection{Chain-of-Thought Reasoning}

Chain-of-thought (CoT) prompting~\cite{wei2022chain} encourages the model to generate intermediate reasoning steps before producing a final answer. Instead of directly outputting a tool call, the model first explains its reasoning: which tool is appropriate, what arguments are needed, and why.

CoT is a form of inference-time compute, trading additional generated tokens for improved accuracy. It is particularly effective for tasks that require multi-step reasoning, such as selecting among many similar tools or constructing complex function arguments. For small models, CoT can compensate for limited implicit reasoning capacity by making the reasoning process explicit.

\section{Retrieval-Augmented Generation}

Retrieval-augmented generation (RAG)~\cite{lewis2020retrieval} augments the model's input with relevant information retrieved from an external knowledge source at inference time. In the context of tool calling, RAG retrieves relevant tool schemas and documentation based on the user's request, rather than including all available tools in every prompt.

This approach offers two benefits for small models. First, it reduces the prompt length by only including relevant tools, which is important given the limited context windows of smaller models. Second, it reduces hallucination by grounding the model's generation in retrieved factual content rather than relying solely on parametric knowledge.

A typical RAG pipeline for tool calling involves encoding tool
descriptions into a vector store, retrieving the top-$k$ most relevant
tools for a given query using semantic similarity, and injecting the
retrieved schemas into the model's prompt before generation.

The retrieval step depends on a dense vector index. Each tool
description is encoded into a fixed-dimensional embedding vector by a
sentence encoder, and all embeddings are stored in an index that
supports approximate nearest-neighbour search. At query time, the
user's request is encoded by the same model, and the $k$ tool
embeddings with the highest cosine similarity to the query embedding
are retrieved. FAISS (Facebook AI Similarity Search) is a widely used
library for this purpose, providing efficient exact and approximate
search over large embedding collections~\cite{johnson2019billion}.

The primary failure mode of RAG for tool calling is candidate
disambiguation. When multiple tools have semantically similar
descriptions (for example, \texttt{calculate\_area},
\texttt{calculate\_\allowbreak{}triangle\_\allowbreak{}area}, and
\texttt{calc\_\allowbreak{}area\_\allowbreak{}triangle}), the retrieval step returns all of them
as high-similarity candidates, and the language model must then select
the correct one. This
selection task requires understanding subtle distinctions in tool
semantics that may not be apparent from descriptions alone. The
quality of retrieval is measured by recall@$k$: the fraction of
queries for which the correct tool appears among the $k$ retrieved
candidates. High recall@$k$ is necessary but not sufficient for
accurate tool calling; the model must still identify the correct tool
from the retrieved set.

\section{Parameter-Efficient Fine-Tuning}

Fine-tuning adapts a pre-trained model to a specific task by continuing training on task-specific data. Full fine-tuning updates all model parameters, which is computationally expensive and requires significant memory. Parameter-efficient fine-tuning (PEFT) methods address this by updating only a small subset of parameters while keeping the rest frozen.

\subsection{LoRA}

Low-Rank Adaptation (LoRA)~\cite{hu2022lora} is the most widely used PEFT method. Instead of updating the full weight matrices during fine-tuning, LoRA injects trainable low-rank decomposition matrices alongside the frozen pre-trained weights. For a weight matrix $W \in \mathbb{R}^{d \times k}$, LoRA learns two smaller matrices $A \in \mathbb{R}^{d \times r}$ and $B \in \mathbb{R}^{r \times k}$, where $r \ll \min(d, k)$, such that the adapted weight becomes $W + AB$.

This reduces the number of trainable parameters by orders of magnitude while achieving performance comparable to full fine-tuning on many tasks. For tool calling, LoRA can be used to fine-tune small models on datasets of function-calling examples, teaching them to select tools and generate arguments more accurately.

\subsection{PEFT Variants}

Beyond LoRA, several other PEFT methods exist. QLoRA~\cite{dettmers2023qlora} combines LoRA with quantization, applying low-rank adaptation on top of a 4-bit quantized base model to further reduce memory requirements. Adapter layers insert small trainable modules between frozen Transformer layers. Prefix tuning prepends trainable continuous vectors to the key and value sequences in each attention layer.

This thesis primarily uses LoRA due to its simplicity, strong empirical
performance, and wide support across frameworks and model architectures.

\section{Model Quantization}

Quantization reduces the numerical precision of model weights from
the 16-bit or 32-bit floating-point formats used during training to
lower-bit integer representations at inference time. A weight stored
as a 4-bit integer requires one quarter of the memory of a 16-bit
floating-point value, enabling models that would otherwise require
multi-GPU setups to run on a single consumer GPU. Secondary benefits
include faster weight loading from storage and reduced memory
bandwidth during inference.

Post-training quantization (PTQ) applies the precision reduction
after training is complete, without additional gradient computation.
This contrasts with quantization-aware training (QAT), which
simulates quantization during training and achieves higher accuracy
at the cost of retraining. PTQ is preferred for large language
models because the cost of retraining at scale is prohibitive; the
challenge is to maintain accuracy despite the information loss from
reduced precision.

\subsection{Activation-Aware Weight Quantization}

Naive PTQ at 4-bit precision produces significant accuracy
degradation because not all weights tolerate uniform quantization
equally. Weights connected to channels with large activation
magnitudes have a disproportionate effect on output quality;
quantizing these salient weights to 4 bits introduces large
relative errors where the model is most sensitive.

Activation-aware Weight Quantization (AWQ)~\cite{lin2024awq}
identifies salient channels by examining activation statistics on a
small calibration dataset and applies a per-channel scaling
transformation that reduces the effective quantization error for
high-magnitude channels. The weights are then quantized uniformly
to INT4 after scaling. The transformation is absorbed into the
preceding layer weights, so inference requires no additional
computation beyond standard INT4 matrix multiplication.

For the Qwen~2.5 7B-Instruct model used in this thesis, AWQ INT4
quantization reduces model memory from 14.25~GiB to 5.20~GiB, a
63.5\% reduction, and reduces model loading time from 212~s to 35~s.
The accuracy cost on BFCL~v4 \texttt{simple\_python} is
$-0.5$~percentage points relative to the bfloat16 baseline, within
run-to-run variance. Pre-quantized AWQ INT4 weights are published
on the Hugging Face Hub for all Qwen~2.5 Instruct
sizes~\cite{qwen2024qwen25}, avoiding the need for manual
quantization.

\section{Related Work}

\subsection{Tool Calling and Benchmarks}

Function-calling benchmarks span a spectrum that trades catalogue
size against task depth. The Berkeley Function Calling Leaderboard
(BFCL)~\cite{bfcl2024} is the standardized single-turn benchmark on
which the configurations in this thesis are evaluated; its categories,
metrics, and the specific test splits used are described with the
experimental setup in Section~\ref{sec:eval-framework}. It is the
reference point against which the other benchmarks below are
positioned.

ToolLLM~\cite{qin2024toolllm} extends function-calling evaluation to
over 16,000 real-world APIs from the RapidAPI catalogue, covering
domains including finance, weather, social media, and e-commerce. At
this scale, the model must select among thousands of APIs with
varying documentation quality and compose multi-step tool chains. The
primary metric is execution accuracy, requiring correct runtime
results rather than structural matching. The scope of ToolLLM
reflects a harder generalization challenge than BFCL, as the API
catalog far exceeds what can be memorized from pre-training data.

At the multi-step end of the spectrum, the $\tau$-bench
benchmark~\cite{taubench2024} evaluates end-to-end agentic task
completion in simulated retail and airline customer service domains,
scoring only whether the final database state matches a ground truth
resolution after a sequence of interdependent tool calls. This thesis
uses $\tau$-bench to probe multi-step reliability beyond single-call
accuracy; its task structure, scoring, and the user-simulator
configuration are detailed in Section~\ref{sec:eval-framework}.

\subsection{Fine-Tuned Small Models for Tool Use}

Gorilla~\cite{patil2023gorilla} demonstrated that a LLaMA model
fine-tuned on API documentation can outperform GPT-4 on API calling
tasks. The key finding is that training data relevance can compensate
for parameter count disadvantage when the task requires structured
generation against specific schemas. Gorilla uses a
retrieval-augmented training procedure in which API documentation is
retrieved during training, teaching the model to use retrieved
context rather than rely on memorized API signatures.

The xLAM model family~\cite{zhang2024xlam} extends the fine-tuning
approach with a large-scale training pipeline that draws from diverse
function-calling datasets. The \texttt{xlam-function-calling-60k}
dataset used in this thesis for LoRA fine-tuning was produced by the
APIGen pipeline~\cite{liu2024apigen}, which generates and
automatically verifies function-calling examples across a wide range
of API types and domains. The diversity is intended to improve
generalization beyond the specific schemas seen during training.
Models trained on this dataset are competitive with frontier systems
on standard tool-use benchmarks.

The format of training outputs has a direct effect on how fine-tuning
interacts with constrained decoding at inference time. This
interaction is observed in the present experiments: two LoRA adapters
trained on the same dataset with identical hyperparameters but
different output formats converge to similar final training loss
(0.203 and 0.229 at step 6,750 respectively) yet differ by 7.0
percentage points in BFCL accuracy when paired with the constrained
decoding pipeline. The accuracy difference is attributable entirely
to whether the training output format matches the inference-time
generation target, not to any difference in training quality. This
result extends Gorilla's data-relevance insight from API source
selection to output format alignment.

A recent survey of small language models for agentic
systems~\cite{sharma2025slmagents} identifies Qwen~2.5 7B as one of
the strongest open-weight SLMs for tool use, attributing its performance to instruction tuning on
function-calling examples and the scale of its pre-training corpus.
The survey notes that most existing evaluations test individual
optimization techniques in isolation, and that systematic
combinations of constrained decoding, quantization, and fine-tuning
remain underexplored. This gap directly motivates the cumulative
evaluation design adopted in this thesis.

\subsection{Constrained Generation}

Willard and Louf~\cite{willard2023efficient} formalized constrained
generation using finite-state machines (FSMs) compiled from regular
expressions and JSON schemas. At each decoding step, the current
grammar state determines the set of valid next tokens; tokens outside
this set receive a logit of negative infinity before sampling,
ensuring that the generated sequence conforms to the grammar by
construction. The key contribution is an efficient offline
compilation algorithm that converts JSON schemas into FSMs before
inference begins, so that the per-step cost at generation time
consists only of a state transition lookup.

The Outlines library~\cite{outlines2024} implements the FSM approach
and is the constrained generation backend used in this thesis
via vLLM~\cite{kwon2023vllm}. Two generation modes are relevant to
the two-stage tool-calling pipeline. The \texttt{guided\_choice} mode
restricts output to one string from a predefined list and is used in
Stage~1 to guarantee that the predicted function name exists in the
available schema. The \texttt{guided\_json} mode restricts output to
a JSON object conforming to a provided schema and is used in Stage~2
to guarantee structural validity of the argument object. Both modes
compile their constraint to a token mask at request initialization
time, amortizing the compilation cost across all generation steps for
a given request.

The per-step overhead of FSM-based constrained generation is low when
schemas are compiled ahead of time. Overhead scales with schema
complexity: schemas with deeply nested structures and many optional
fields require larger FSM state spaces and more expensive state
transitions. For the function schemas in BFCL \texttt{simple\_python},
complexity is moderate and the overhead falls within practical
inference budgets. Small models benefit disproportionately from
constrained generation because their baseline format error rates are
higher: without constraints, Qwen~2.5 7B produces unparseable output
on 98.5\% of BFCL test cases, a failure rate that constrained
decoding eliminates entirely.

LMQL~\cite{beurer2023lmql} offers an alternative approach through
a declarative query language in which output constraints are expressed
as Python predicates evaluated during generation. The approach is
more flexible than FSM-based methods for constraints that are
difficult to specify as formal grammars. The per-step cost is higher,
however, because predicates are evaluated at each generation step
rather than looked up from a precompiled state table. This thesis
uses FSM-based constrained decoding via vLLM because it is better
integrated with batch inference and provides formal guarantees from a
compiled grammar state.

\subsection{Cascade Architectures}

A cascade architecture routes inference requests across models of
different capability and cost. Routine
requests are handled by a smaller, faster model; requests that exceed
its reliable operating range are escalated to a larger model,
typically accessed through an external API. Routing decisions may be
based on output confidence, output validation, or a query complexity
estimate computed before inference.

For tool-calling tasks, structural validation is a natural routing
gate. If the small model output fails to parse as valid JSON or names
a function not present in the available schema, the response can be
rejected and escalated without execution. Constrained decoding removes
this failure mode by construction, since every output is
structurally valid by guarantee. After constrained decoding is
applied, the residual failure rate is attributable entirely to
incorrect argument values, which structural checks cannot detect.
This shifts the routing problem from format detection to semantic
verification, a harder problem that requires either ground-truth
access or execution feedback.

The residual semantic failure rate determines the fraction of traffic
that must be escalated to a frontier model and the associated
variable inference cost. CD achieves 72.75\% accuracy,
implying a 27.25\% escalation rate under structural routing. Config
CD+FT-aligned reduces this to 23.25\%, a reduction of approximately
4.0 percentage points. The relationship between optimization
technique selection and cascade routing cost is a practical design
consideration that has received limited attention in prior work; this
thesis provides direct measurements for five technique combinations.

\subsection{Small Model Optimization}

Recent model families have narrowed the capability gap with frontier
models through improvements applied at training time. Phi-4~%
\cite{abdin2024phi} achieves competitive performance with models
several times its size on reasoning and mathematics benchmarks,
demonstrating that training data curation can compensate for reduced
parameter count. Llama~3~\cite{llama2024llama32} introduces
grouped-query attention and an expanded vocabulary to improve
inference efficiency and multilingual coverage.
Qwen~2.5~\cite{qwen2024qwen25} achieves strong tool-use performance
through instruction tuning on function-calling examples and a context
window of 128,000 tokens.

Knowledge distillation~\cite{hinton2015distilling} provides a
complementary path to capable small models. A small student model is
trained to match the output distribution of a large teacher rather
than ground-truth labels, transferring reasoning capabilities that
the student would not acquire from labeled data alone. Unlike the
post-training techniques evaluated in this thesis, distillation
affects the model's parametric knowledge and requires access to the
teacher model's outputs during training.

Quantization reduces model memory at the cost of a small accuracy
penalty. For the Qwen~2.5 7B model used in this thesis, AWQ INT4
quantization~\cite{lin2024awq} reduces memory from 14.25~GiB to
5.20~GiB and reduces accuracy by 0.5 percentage points on BFCL
\texttt{simple\_python}, a cost within run-to-run variance. This
result is consistent with reported accuracy retention for AWQ across
other model families at similar parameter scales.

The pre-training and instruction-tuning improvements embodied in
Phi-4, Qwen~2.5, and Llama~3 are largely orthogonal to the
inference-time and post-training techniques evaluated here.
Constrained decoding, quantization, and LoRA fine-tuning can be
applied on top of any pre-trained checkpoint. The question addressed
in this thesis is how much additional accuracy is recovered by
combining these techniques on top of an already well-trained small
model, and which combinations offer the best accuracy-cost tradeoff
for deployment in a cascade architecture.
