\chapter{Methodology} \label{ch:methodology}

This chapter describes how the optimization techniques are evaluated. The
experiments follow a cumulative ladder: a baseline configuration is established
first, and one technique is added at each subsequent step to isolate its
marginal contribution. The Qwen~2.5 model family is evaluated across four
sizes from 0.5B to 7B. The techniques covered are constrained decoding, schema
enrichment, quantization, few-shot prompting, chain-of-thought prompting,
retrieval-augmented generation, and LoRA fine-tuning. All configurations are
evaluated on the Berkeley Function Calling Leaderboard.

\section{Experimental Design}

The experimental design is based on controlled comparisons and runs in two
complementary directions. In the first, a baseline configuration is established
and each technique is added one at a time, which measures the marginal gain of
each technique on top of the existing stack. In the second, the inference-time
techniques are evaluated without constrained decoding, which measures what each
technique achieves on its own and quantifies the contribution of constrained
decoding itself.

\subsection{Cumulative Evaluation Ladder}

Twelve configurations are evaluated in total. Eleven form the cumulative
ladder in Figure~\ref{fig:eval-ladder}; the twelfth, CD+schema, branches off
separately and is reported in Section~\ref{sec:cd-schema}.

The ladder is organised around a main chain (Config~B $\to$ CD $\to$ CD+Q
$\to$ CD+Q+FT-aligned) that starts from an unconstrained full-precision
baseline and adds one technique at each step: constrained decoding, then
AWQ INT4 quantization, then a format-aligned LoRA adapter. The remaining
configurations branch from this chain to isolate a single design choice
rather than to extend the stack. Few-shot prompting (PE) branches from CD at
full precision, so its effect is separated from quantization; chain-of-thought
(ITC) and retrieval (RAG) branch from CD+Q; and two LoRA adapters, misaligned
(CD+FT) and format-aligned (CD+FT-aligned), branch from CD\@. Two further
variants drop constrained decoding entirely (FT-only and FT-aligned-ng, where
ng denotes no guided decoding) to measure fine-tuning on its own.

The second arm of the design re-evaluates the three inference-time techniques
without guided decoding (Config~PE-ng, ITC-ng, and RAG-ng) to quantify the
contribution of constrained decoding itself. These isolation variants are
reported in Section~\ref{sec:technique-isolation}; they are not rungs on the
ladder and are omitted from Figure~\ref{fig:eval-ladder}.

\begin{figure}[ht]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
    box/.style={
        rectangle, draw, rounded corners=4pt,
        minimum width=2.8cm, minimum height=0.85cm,
        align=center, font=\small
    },
    main/.style={box, fill=blue!8},
    side/.style={box, fill=orange!8},
    para/.style={box, fill=gray!10, draw=gray!60},
    ft/.style={box, fill=green!8},
    arr/.style={-{Stealth}, thick},
    darr/.style={-{Stealth}, thick, dashed, gray!60}
]

% Main chain (top row)
\node[main] (B)        {Base};
\node[main, right=1.5cm of B]   (CD)       {CD};
\node[main, right=1.5cm of CD]  (CDQ)      {CD+Q};
\node[main, right=1.5cm of CDQ] (CDQFTa)   {CD+Q+\\FT-aligned};

% Inference-time branches from CD+Q (second row)
\node[side, below=1.1cm of CDQ, xshift=-1.7cm] (CDQITC) {CD+Q+ITC};
\node[side, below=1.1cm of CDQ, xshift= 1.7cm] (CDQRAG) {CD+Q+RAG};

% Few-shot branch from CD (second row, full precision)
\node[side, below=1.1cm of B, xshift=0.8cm] (PE) {PE};

% Bottom row: unguided standalone (gray) and fine-tuning branches (green),
% chained left to right so the boxes cannot overlap
\node[para, below=2.4cm of B, xshift=-3.8cm] (FTonly) {FT-only};
\node[para, right=0.4cm of FTonly] (FTalng) {FT-aligned-ng};
\node[ft,   right=0.4cm of FTalng] (CDFT)   {CD+FT};
\node[ft,   right=0.4cm of CDFT]   (CDFTa)  {CD+FT-aligned};

% Main chain arrows
\draw[arr] (B)   -- node[above, font=\footnotesize]{$+$CD}           (CD);
\draw[arr] (CD)  -- node[above, font=\footnotesize]{$+$Q}            (CDQ);
\draw[arr] (CDQ) -- node[above, font=\footnotesize]{$+$FT\textsubscript{aligned}} (CDQFTa);

% Inference-time tracks from CD+Q
\draw[arr] (CDQ.south) -- node[left,  font=\footnotesize]{$+$ITC} (CDQITC.north);
\draw[arr] (CDQ.south) -- node[right, font=\footnotesize]{$+$RAG} (CDQRAG.north);

% Fine-tuning tracks from CD
\draw[darr] (CD.south) -- node[left, font=\footnotesize, fill=white, inner sep=1pt, pos=0.3]{$+$FT} (CDFT.north);
\draw[darr] (CD.south) -- node[left, font=\footnotesize, fill=white, inner sep=1pt, pos=0.65]{$+$FT\textsubscript{aligned}} (CDFTa.north);

% Few-shot track from CD
\draw[arr] (CD.south) -- node[left, font=\footnotesize, fill=white, inner sep=1pt]{$+$few-shot} (PE.north);

% Unguided standalone tracks from B
\draw[darr] (B.south) -- (FTonly.north);
\draw[darr] (B.south) -- (FTalng.north);

\end{tikzpicture}%
}
\caption{Incremental evaluation ladder showing eleven of the twelve
configurations. The main chain (top row, solid arrows) adds one technique at a
time. Orange nodes are inference-time branches: PE branches from CD at full
precision; ITC and RAG branch from CD+Q\@.
Green nodes are fine-tuning branches from CD\@.
Gray nodes are unguided standalone configurations (no constrained decoding).
Not shown: the off-ladder CD+schema variant (Section~\ref{sec:cd-schema}) and
the no-guided isolation variants PE-ng, ITC-ng, and RAG-ng
(Section~\ref{sec:technique-isolation}), which are evaluated separately.}
\label{fig:eval-ladder}
\end{figure}

\subsection{Hardware and Infrastructure}

All experiments are conducted on the DTU HPC cluster
(gbar.dtu.dk), running AlmaLinux~9.7 under the LSF job
scheduler. Each job is allocated one NVIDIA L40S 48\,GB
GPU in exclusive process mode; inference jobs use four CPU
cores and 32\,GB of RAM, while LoRA training jobs use eight
cores and 64\,GB. Inference jobs have a two-hour wall-time
limit and training jobs an eight-hour limit.

Model inference is served via vLLM~0.8.5, which exposes a
REST API on localhost that the evaluation client connects to,
with GPU memory utilisation limited to 90\,\%. Full-precision models are loaded in bfloat16 (14\,GB
VRAM for the 7B model); quantized models are loaded in float16
under the AWQ Marlin backend (4--5\,GB VRAM). The software
environment uses uv~0.5.13, Python~3.12.11, and CUDA~12.6.3.
Model weights are cached from the Hugging Face Hub on cluster
storage.

The evaluation pipeline, prompt templates, guided decoding schemas,
HPC job scripts, and training scripts are available in the
accompanying repository.\footnote{\url{https://github.com/pabeckha/slm-agents}}

\section{Evaluation Framework} \label{sec:eval-framework}

Two benchmarks are used to evaluate the configurations.
The Berkeley Function Calling Leaderboard covers single-turn
tool-calling accuracy across categories of increasing
complexity. $\tau$-bench covers multi-step agentic task
completion, where the model must reason, call tools, observe
results, and iterate until a task is done.

\subsection{Berkeley Function Calling Leaderboard}

All configurations are evaluated on the Berkeley Function
Calling Leaderboard (BFCL) v4~\cite{bfcl2024}. Four
categories are used: \texttt{simple\_python} (400 test
cases, one candidate function per test), \texttt{multiple}
(the model must select the correct function from several
candidates), \texttt{parallel} (the model must produce
multiple simultaneous calls to the same function), and
\texttt{parallel\_multiple} (simultaneous calls to
different candidate functions). Each
test case consists of a user prompt and one or more function
definitions; the model is required to produce the correct
function call with the correct arguments. Frontier model
scores are taken from the published BFCL leaderboard rather
than re-evaluated, as the official scores are the canonical
reference for the benchmark.

The primary metric is AST accuracy: the model output is
parsed and compared structurally against the ground truth,
checking function name, argument names, argument values, and
argument types. BFCL also defines an execution accuracy
metric, which runs the generated call and compares its
runtime result against a reference output; AST accuracy is
used throughout this thesis because it is deterministic,
reproducible, and independent of the runtime environment,
consistent with the leaderboard convention for comparing
configurations systematically. Evaluation is performed using the BFCL
\texttt{ast\_checker}, which applies lenient type coercion
between compatible types (e.g., integer and float). All
inference runs use temperature 0 to ensure deterministic
outputs across configurations; at temperature~0 vLLM applies
greedy decoding, producing identical outputs on repeated
runs within a fixed software environment. The CD+Q result
(289/400) was confirmed across six consecutive re-runs that
produced identical predictions. Re-runs spread over the
project period, however, varied by one to two cases out of
400 (CD scored between 289 and 291, and CD+Q between 288 and
289, across runs months apart), so any single accuracy
figure carries a run-to-run uncertainty of approximately
$\pm$0.5~pp. Headline figures quote the originally recorded
runs; the paired statistical tests in
Section~\ref{sec:combined-configs} are computed on the
stored per-case prediction files. Function selection accuracy
and argument extraction accuracy are reported separately to
identify where failures occur along the prediction pipeline.

To measure per-call inference latency, a sequential latency run
(parallel dispatch disabled) is conducted on the first 100
\texttt{simple\_python} test cases for three configurations: B,
CD, and CD+Q. Sequential dispatch ensures that each measurement
is a clean per-call latency without parallelism overlap. The same
vLLM server and hardware are used as for the accuracy runs.
Mean, median, and 95th-percentile latency are reported.

\subsection{$\tau$-bench}

The $\tau$-bench retail benchmark~\cite{taubench2024} evaluates
end-to-end multi-step task completion. Each task is a simulated
customer service interaction in which the model must reason across
5--15 tool calls, observe intermediate results, and reach a final
database state that matches a ground truth resolution. CD and
CD+FT-aligned are evaluated on the retail domain test split (115 tasks);
the remaining configurations are not evaluated on $\tau$-bench because
the primary bottleneck identified is multi-turn planning rather than
single-call argument accuracy.

Three full evaluation runs are completed to account for variance
introduced by the user simulator. The user simulator is Qwen~2.5
7B-Instruct served on the same vLLM instance as the agent model,
running at temperature~0.3 to produce naturalistic user responses.
The agent model runs at temperature~0.0 to ensure deterministic
tool call generation. All runs are executed on a single NVIDIA L40S
GPU (48~GiB) on the DTU HPC cluster. This simulator choice departs
from the reference $\tau$-bench setup, which drives the user
simulator with a frontier model; a 7B simulator may misstate or omit
task details, which deflates pass rates independently of agent
capability. The implications are discussed in the threats to
validity (Section~\ref{sec:threats-validity}).

Scoring is binary: a task receives reward~1.0 only if the final
database state exactly matches the ground truth at the end of the
interaction, and reward~0.0 otherwise. Partial task completion
yields no reward. The pass rate is reported as the mean reward
across all tasks and all runs.

\section{Models}
\subsection{Qwen 2.5}

The Qwen 2.5 model family~\cite{qwen2024qwen25} is used
across all experiments. Four sizes are evaluated: 0.5B,
1.5B, 3B, and 7B parameters, all in the Instruct variant.
The 7B model is the primary subject of analysis; the
smaller sizes are included to study how model scale
interacts with each optimization technique.

Qwen 2.5 is selected for three reasons. First, the family
covers a continuous size range under a single architecture
and training regime, which allows scale effects to be
isolated without confounding differences in model design.
Second, native AWQ INT4 quantized variants are available
for all sizes on the Hugging Face Hub, avoiding the need
for manual quantization. Third, recent surveys of SLMs for
agentic systems identify Qwen 2.5 7B as one of the
strongest open-weight models for tool use and function
calling at this parameter scale~\cite{sharma2025slmagents}.
All models are served under an Apache 2.0 licence.

All Qwen~2.5 models share a decoder-only Transformer
architecture with grouped-query attention (GQA), SwiGLU
feed-forward activations, and rotary position embeddings
(RoPE). The 7B variant uses 28 layers, a hidden dimension
of 3584, and 28 query heads grouped into 4 key-value head
groups. The supported context length is 128{,}000 tokens,
though all experiments in this thesis operate well within
this limit. GQA reduces the KV cache memory footprint
relative to standard multi-head attention, which is one
reason the 7B model fits comfortably on a single consumer
GPU in bfloat16 precision.

The Instruct variant is pre-trained on a multilingual corpus of approximately
18 trillion tokens. Supervised fine-tuning on instruction-following and tool-use
examples, followed by alignment via reinforcement learning from human feedback,
produces a model that follows instructions reliably. It responds to structured
prompts and tool schemas without additional task-specific training, serving as
a practical starting point for the optimization ladder evaluated in this thesis.

\subsection{Contrast Checkpoints}
\label{sec:contrast-checkpoints}

The cross-family checks in Chapter~\ref{ch:results} use four additional instruct
models from three other families: Llama~3.2 1B-Instruct and Llama~3.2
3B-Instruct~\cite{metaai2024llama32}, Phi-4-mini-Instruct~\cite{abdin2024phi},
and Gemma~3 1B-Instruct~\cite{gemmateam2025gemma3}. These are contrast
checkpoints rather than part of the primary size ladder. Their purpose is to
test whether the findings established on Qwen~2.5 depend on one model's training
or hold under different pre-training corpora and instruction-tuning recipes.

The four models span roughly 1B to 4B parameters: Gemma~3 1B and Llama~3.2 1B
sit near 1B, Llama~3.2 3B at 3.2B, and Phi-4-mini at 3.8B, overlapping the lower
half of the Qwen range. Their training regimes differ in ways that matter for
this check. Phi-4 emphasises curated and synthetic training data, the small
Llama~3.2 models are distilled from larger Llama checkpoints, and Gemma~3 is
likewise trained with distillation. A result that survives all four is unlikely
to be an artefact of any single training pipeline.

The contrast checkpoints enter the evaluation in three places. The
constrained-decoding generalization check (Section~\ref{sec:cross-family-cd})
evaluates Config~CD on BFCL~v4 \texttt{simple\_python} for each checkpoint, to
test whether the structural benefit established on Qwen holds under different
pre-training corpora. The format-aligned fine-tuning analysis
(Section~\ref{sec:format-aligned}) trains a format-aligned adapter on each model
and evaluates it without guided decoding, to check whether the standalone
degradation seen on Qwen also appears across families. The GitHub MCP evaluation
(Section~\ref{sec:github-mcp}) compares baseline and constrained tool selection
across all five families on the 21-tool catalogue. The quantized, fine-tuned,
and $\tau$-bench configurations stay on Qwen~2.5; extending them across the
contrast families is left to future work. Llama~3.2 is served under
the Llama~3.2 Community License, Phi-4-mini under the MIT License, and Gemma~3
under the Gemma Terms of Use.

\section{Baseline Definition}
\label{sec:baseline-def}

Base is the baseline configuration: the model is prompted
without constrained decoding, at full precision (bfloat16),
with no quantization, no fine-tuning, and no few-shot
examples. The pipeline is two-stage (Figure~\ref{fig:pipeline}). In the first stage,
the available function signatures and descriptions are
presented to the model, and the prompt ends with
\texttt{Function name:~}; the model completes freely.
In the second stage, the selected function signature and
the user query are presented, and the prompt ends with
\texttt{JSON:~}; the model generates argument values as
free-form text. A JSON extraction heuristic is applied to
the output, capturing the substring from the first
opening brace to the last closing brace. Temperature is set
to 0 across all configurations to ensure deterministic
outputs, removing randomness as a confound when comparing
techniques.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance=0.8cm,
    inp/.style={rectangle, draw=gray!60, fill=gray!10, rounded corners=3pt,
                minimum width=2.6cm, minimum height=0.72cm,
                align=center, font=\small},
    stg/.style={rectangle, draw, fill=blue!8, rounded corners=4pt,
                minimum width=5.2cm, minimum height=1.05cm,
                align=center, font=\small},
    res/.style={rectangle, draw, fill=green!10, rounded corners=3pt,
                minimum width=2.6cm, minimum height=0.72cm,
                align=center, font=\small},
    arr/.style={-{Stealth}, thick},
]

\node[stg] (s1)
    {Stage~1: Function Selection\\[3pt]{\footnotesize\texttt{guided\_choice}}};

\node[inp, above=0.8cm of s1, xshift=-1.6cm] (q1)   {User Query};
\node[inp, above=0.8cm of s1, xshift= 1.6cm] (sigs) {All Function\\Signatures};

\node[res, below=0.8cm of s1] (fn) {Function Name};

\node[stg, below=0.8cm of fn] (s2)
    {Stage~2: Argument Extraction\\[3pt]{\footnotesize\texttt{guided\_json}}};

\node[inp, left=0.8cm of s2] (q2) {User Query};

\node[res, below=0.8cm of s2] (js) {JSON Arguments};

\draw[arr] (q1.south)   -- (s1.north);
\draw[arr] (sigs.south) -- (s1.north);
\draw[arr] (s1)         -- (fn);
\draw[arr] (fn)         -- (s2);
\draw[arr] (q2)         -- (s2);
\draw[arr] (s2)         -- (js);

\end{tikzpicture}
\caption{Two-stage inference pipeline. Stage~1 selects a function name from
all available candidates; Stage~2 extracts argument values for the selected
function. In CD and above, \texttt{guided\_choice} and
\texttt{guided\_json} enforce structural validity at each stage.}
\label{fig:pipeline}
\end{figure}

Base establishes the lower bound for the evaluation ladder: the model is
presented with the generic, model-agnostic pipeline and no constrained
decoding, so any improvement at a later rung is attributable to the
technique added there. Its accuracy and failure breakdown are reported in
Chapter~\ref{ch:results}.

The argument-extraction stage parses the completion as JSON under two
criteria. The headline baseline applies a strict whole-completion parse: the
entire model output must load as a single JSON object, which is the floor a
naive model-agnostic integration achieves (1.50\%). The technique-isolation
arm (Section~\ref{sec:technique-isolation}) instead reads the first complete
JSON object and ignores trailing text, so the prompt-level techniques are
compared fairly; under this lenient criterion the same Base run scores
62.00\%. An earlier no-guided parser sliced from the first to the last brace
and so failed on a valid answer followed by further text; it was replaced by
the first-object parser used for all corrected no-guided figures.

Base is deliberately model-agnostic: it does not use the
tool-calling prompt template that Qwen~2.5 provides. The 1.5\% figure
measures what a generic integration achieves without model-specific
prompting, which is the relevant floor for the technique ladder
evaluated here: every subsequent configuration changes exactly one
component relative to this fixed pipeline.

A control configuration, B-template, bounds this design choice from
above. It prompts the same model through its native tool-calling
chat template, passing the full unmodified BFCL function schemas
(including parameter descriptions, default values, and
enumerations), and parses the model's native \texttt{<tool\_call>}
output with no guided decoding. Two properties separate it from Base:
the prompt format matches the model's instruction tuning, and the
full schema detail is visible to the model, whereas the
model-agnostic pipeline carries only parameter names and types. In
exchange, B-template provides no structural validity guarantee and
applies only to models that ship a tool-calling template. Its accuracy
is reported in Chapter~\ref{ch:results}, and the implications of this
control for interpreting the residual failure rate of the constrained
configurations are discussed in Chapter~\ref{ch:discussion}.

Figure~\ref{fig:base-vs-template} makes the two designs concrete for the
\texttt{add} query from Chapter~\ref{ch:introduction}. The model-agnostic
pipeline (a) anchors generation with the plain-text \texttt{JSON:~} cue and
reduces each parameter to a \texttt{name (type)} pair, so the model receives
neither the format it was instruction-tuned on nor the parameter descriptions,
defaults, and enumerations present in the source schema. The native template
(b) renders the same request the way the model saw tool use during training:
the full JSON schema is placed inside a \texttt{<tools>} block in the system
message, and the model is asked to wrap its call in \texttt{<tool\_call>} tags,
which the backend then parses. The contrast isolates the two confounds folded
into the Base score, prompt format and schema completeness, from the model's
underlying capability.

\begin{figure}[ht]
\centering
\lstdefinestyle{promptbox}{%
    basicstyle=\ttfamily\scriptsize,
    frame=single,
    framesep=4pt,
    columns=fullflexible,
    keepspaces=true,
    breaklines=true,
    breakatwhitespace=false,
    postbreak=\mbox{\textcolor{gray}{$\hookrightarrow$}\space},
    showstringspaces=false,
    aboveskip=4pt,
    belowskip=2pt,
}

\textbf{(a) Base --- generic, model-agnostic pipeline (Stage~2 prompt)}
\begin{lstlisting}[style=promptbox]
Function: add(a: number, b: number) -> number
Description: Add two numbers
User request: What is the sum of 265 and 345?
Extract the argument values as a JSON object with keys: a (number), b (number).
JSON:
\end{lstlisting}
\textit{Model output (free generation; strict whole-completion parse fails):}
\begin{lstlisting}[style=promptbox]
{"a": 265, "b": 345} The function `add` takes two numbers as arguments
and returns their sum [...] So, the sum of 265 and 345 is 610.
\end{lstlisting}

\vspace{4pt}
\textbf{(b) B-template --- native tool-calling chat template (rendered prompt)}
\begin{lstlisting}[style=promptbox]
<|im_start|>system
You are Qwen, created by Alibaba Cloud. You are a helpful assistant.

# Tools

You may call one or more functions to assist with the user query.

You are provided with function signatures within <tools></tools> XML tags:
<tools>
{"type": "function", "function": {"name": "add", "description": "Add two
numbers", "parameters": {"type": "object", "properties": {"a": {"type":
"number"}, "b": {"type": "number"}}, "required": ["a", "b"]}}}
</tools>

For each function call, return a json object with function name and arguments
within <tool_call></tool_call> XML tags:
<tool_call>
{"name": <function-name>, "arguments": <args-json-object>}
</tool_call><|im_end|>
<|im_start|>user
What is the sum of 265 and 345?<|im_end|>
<|im_start|>assistant
\end{lstlisting}
\textit{Model output (parsed from the \texttt{<tool\_call>} block):}
\begin{lstlisting}[style=promptbox]
<tool_call>
{"name": "add", "arguments": {"a": 265, "b": 345}}
</tool_call><|im_end|>
\end{lstlisting}

\caption{The same \texttt{add} request under the two baseline integrations.
\textbf{(a)} The model-agnostic Base pipeline uses plain-text anchors
(\texttt{JSON:}) and a reduced \texttt{name (type)} parameter listing; the
model continues past the leading JSON object with explanatory prose, so a
strict whole-completion parse rejects the output (1.50\,\%). \textbf{(b)}
B-template renders the request through Qwen~2.5's native chat template: the
full schema is supplied inside \texttt{<tools>}, ChatML control tokens
(\texttt{<|im\_start|>}, \texttt{<|im\_end|>}) mark turns, and the model emits
its call inside \texttt{<tool\_call>} tags, which the backend extracts
(96.00\,\%). The two panels differ in prompt format and in how much of the
source schema reaches the model; both are held fixed and replaced by a
structural guarantee in CD.}
\label{fig:base-vs-template}
\end{figure}

\section{Constrained Decoding}

CD adds constrained decoding to Base; all other
settings remain unchanged. Constrained decoding is
implemented via vLLM's guided decoding
feature~\cite{kwon2023vllm, willard2023efficient}, which
applies logit masking at each generation step to restrict
the output to a valid token sequence.

Two mechanisms are used in the two-stage pipeline. In the
function selection stage, \texttt{guided\_choice} restricts
the output to one of the valid function names, guaranteeing
that the selected name exists in the available function
set. In the argument extraction stage, \texttt{guided\_json}
restricts the output to a JSON object that conforms to a
schema derived from the function definition, with parameter
names, types, and required fields enforced, and
\texttt{additionalProperties} set to false.

For the \texttt{parallel} and \texttt{parallel\_multiple} categories,
which require several simultaneous calls, the single-name selection
stage is replaced. The initial design generated a JSON object of the
form \texttt{\{"calls": [\dots]\}} under \texttt{guided\_json}, where the
array items were restricted to the set of candidate function names and
the maximum array length equalled the number of distinct candidates;
each selected name was then passed through the standard argument
extraction stage. This design emits at most one argument object per
distinct function name: because argument extraction is conditioned only
on the user query and the selected function, and decoding is
deterministic at temperature~0, two calls to the same function would
receive identical argument objects. Every \texttt{parallel} case
provides exactly one distinct candidate function, so this design
produces a single call by construction.

To support repeated calls to the same function, the schema is extended
to a joint multi-call form: rather than selecting a list of names and
then extracting arguments per name, the model generates a JSON array of
complete call objects in a single guided-decoding pass, where each
object carries its own function name and argument bindings and the same
function may recur with distinct arguments. Both schemas are evaluated.
The single-extraction design defines the boundary of the per-function
pipeline, and the joint multi-call schema is the remedy for it; the
consequences of each for the \texttt{parallel} category
are analysed in Section~\ref{sec:bfcl-extended} and
Chapter~\ref{ch:discussion}.

Figure~\ref{fig:single-vs-joint} contrasts the two schemas. The
single-extraction design selects a list of distinct function names and
then extracts arguments for each name independently, so it emits at most
one argument object per distinct name. The joint multi-call schema instead
produces a single JSON array of complete \texttt{\{name, arguments\}}
objects, in which the same function may appear more than once with different
arguments.

\begin{figure}[ht]
\centering
\textbf{(a) Single-extraction schema (per-function pipeline)}\\[2pt]
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
    node distance=0.9cm,
    inp/.style={rectangle, draw=gray!60, fill=gray!10, rounded corners=3pt,
                minimum width=2.2cm, minimum height=0.8cm, align=center, font=\small},
    stg/.style={rectangle, draw, fill=blue!8, rounded corners=4pt,
                minimum width=3.4cm, minimum height=1.0cm, align=center, font=\small},
    res/.style={rectangle, draw, fill=green!10, rounded corners=3pt,
                minimum width=2.8cm, minimum height=0.8cm, align=center, font=\small},
    arr/.style={-{Stealth}, thick},
]
\node[inp] (q) {User query};
\node[stg, right=of q] (s1) {Stage 1: \texttt{guided\_json}\\select name list\\$\{f_1, f_2\}$};
\node[stg, right=of s1] (s2) {Stage 2: \texttt{guided\_json}\\extract args\\per distinct name};
\node[res, right=of s2] (out) {$f_1(\mathrm{args}_1)$\\$f_2(\mathrm{args}_2)$};
\draw[arr] (q) -- (s1);
\draw[arr] (s1) -- (s2);
\draw[arr] (s2) -- (out);
\node[font=\footnotesize\itshape, text width=5.2cm, align=center, below=0.35cm of s2]
    {A repeated call to $f_1$ collapses: one argument object per distinct name.};
\end{tikzpicture}%
}

\vspace{8pt}
\textbf{(b) Joint multi-call schema (single pass)}\\[2pt]
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
    node distance=0.9cm,
    inp/.style={rectangle, draw=gray!60, fill=gray!10, rounded corners=3pt,
                minimum width=2.2cm, minimum height=0.8cm, align=center, font=\small},
    stg/.style={rectangle, draw, fill=blue!8, rounded corners=4pt,
                minimum width=5.0cm, minimum height=1.0cm, align=center, font=\small},
    res/.style={rectangle, draw, fill=green!10, rounded corners=3pt,
                minimum width=3.0cm, minimum height=0.8cm, align=center, font=\small},
    arr/.style={-{Stealth}, thick},
]
\node[inp] (q2) {User query};
\node[stg, right=of q2] (sj) {Single \texttt{guided\_json} pass\\JSON array of \{name, arguments\} objects};
\node[res, right=of sj] (out2) {$f_1(\mathrm{args}_1)$\\$f_1(\mathrm{args}_2)$\\$f_2(\mathrm{args}_3)$};
\draw[arr] (q2) -- (sj);
\draw[arr] (sj) -- (out2);
\node[font=\footnotesize\itshape, text width=5.2cm, align=center, below=0.35cm of sj]
    {Each object carries its own name and arguments; the same function may recur.};
\end{tikzpicture}%
}
\caption{Single-extraction versus joint multi-call constrained-decoding
schemas for the \texttt{parallel} categories. \textbf{(a)} The
single-extraction pipeline first selects a list of distinct function names,
then extracts an argument object for each name in a separate
\texttt{guided\_json} pass; because argument extraction is conditioned only
on the query and the selected name and decoding is deterministic at
temperature~0, two calls to the same function would be identical, so the
design emits one object per distinct name. \textbf{(b)} The joint multi-call
schema generates a single JSON array of complete call objects in one pass,
so the same function can appear multiple times with different arguments.
This is the change that lifts \texttt{parallel} from a structural $0.0\%$ to
$79.0\%$ at 7B (Section~\ref{sec:bfcl-extended}).}
\label{fig:single-vs-joint}
\end{figure}

The resulting accuracy for CD, together with the function-selection rate
and the breakdown of residual argument-value failures, is reported in
Chapter~\ref{ch:results}.

\section{Quantization}

CD+Q adds AWQ INT4 quantization~\cite{lin2024awq}
to CD. The pre-quantized model
\texttt{Qwen/\allowbreak{}Qwen2.5-\allowbreak{}7B-\allowbreak{}Instruct-\allowbreak{}AWQ} is used, loaded via
vLLM with the \texttt{awq\_marlin} backend, which applies
optimized Marlin kernels for efficient INT4 inference on
GPU. All other settings remain unchanged.

AWQ is selected because activation-aware weight quantization
preserves accuracy better than uniform quantization by
protecting weights that correspond to large activation
values. Pre-quantized variants for all Qwen~2.5 sizes are
available on the Hugging Face Hub, avoiding the need for
a local quantization step.

CD+Q achieves 72.25\% AST accuracy (289/400), a reduction
of 0.5~pp relative to CD, within run-to-run variance.
AWQ INT4 quantization reduces the 7B model's GPU memory
footprint from 14.25~GiB to 5.20~GiB (63.5\% reduction)
and startup time from 212~s to 35~s. No new failure modes
are introduced: the failure distribution remains semantic,
with the same categories of argument value errors as CD.

\section{Prompt Engineering and Few-Shot}

PE extends CD by prepending three few-shot examples to the
argument extraction prompt. The function selection prompt is left unchanged;
examples are added only to the argument extraction stage, where all CD
failures occur.

The three examples are constructed to target the most frequent failure
categories observed in CD by inspecting the full set of 109 incorrect
predictions on the test split. The examples are not drawn from the test
split itself; they are hand-authored to illustrate the identified
failure classes. This design choice is a limitation: the failure
categories are identified from the same test split used to evaluate PE,
so the examples are informed by test-set error patterns rather than a
held-out development set. The first example demonstrates numeric
precision: the expected value is \texttt{9.81}, not the rounded approximation
\texttt{9.8}. The second demonstrates string format with a location qualifier:
the expected value is \texttt{"San Francisco, CA"}, not
\texttt{"San Francisco"}. The third demonstrates Python expression syntax: the
expected value is \texttt{x**2}, not the mathematical notation
\texttt{x\textasciicircum{}2}. Three examples are selected as the minimum number needed to
cover distinct failure classes; additional examples would increase context
length without introducing new failure categories.

All other settings remain unchanged from CD: the full-precision
Qwen~2.5 7B-Instruct model is used, guided decoding is applied to
both pipeline stages, and temperature is set to~0. PE is evaluated
as a parallel track to CD+Q rather than as an extension of it, so the
full-precision model is retained to isolate the effect of the few-shot
examples from any interaction with quantization.

\section{Inference-Time Compute}
\label{sec:method-itc}

CD+Q+ITC extends CD+Q with a chain-of-thought reasoning
step~\cite{wei2022chain,snell2024scaling}. The two-stage pipeline is
unchanged at its endpoints: function selection still uses
\texttt{guided\_choice} and argument extraction still uses
\texttt{guided\_json}. A free-form reasoning stage is inserted between
them. After the function name is selected, the model generates a reasoning
trace of up to 256 tokens, prompted to set out which arguments the call
requires and what values they should take, before the final guided
extraction.

The reasoning stage is deliberately unconstrained. The structural guarantee
of constrained decoding applies only to the final extraction, where it is
needed to produce a valid call; the intermediate trace is left free so the
model can deliberate about argument values without the token mask
restricting it. The 256-token budget bounds the additional inference cost
while leaving room for the model to reason over the argument set of a
single call. The reasoning text is discarded after the extraction step and
does not enter the parsed output.

CD+Q+ITC branches from CD+Q rather than extending the stack, so the
AWQ~INT4 model is retained and the only change relative to CD+Q is the
inserted reasoning stage. All other settings are inherited from CD+Q:
guided decoding on both endpoints and temperature~0. Isolating the effect
of inference-time compute from quantization at the 7B scale, where the
quantization penalty is 0.5~pp, is not required because the reasoning stage
is evaluated as a single addition to the CD+Q operating baseline.

\section{Retrieval-Augmented Generation}

CD+Q+RAG extends CD+Q by replacing the per-test-case oracle
function provision with a retrieval step~\cite{lewis2020retrieval}. In all
prior configurations, each test case is presented with the exact function
definition required to answer it. In CD+Q+RAG, the model receives
instead the top-$k$ functions retrieved from an index of the full function
catalogue.

All 370 unique function signatures from the \texttt{simple\_python} category
are embedded using \texttt{sentence-transformers/all-MiniLM-L6-v2} and indexed
with FAISS before evaluation begins. The \texttt{all-MiniLM-L6-v2} model is
selected because it is designed for semantic similarity of short texts, requires
no GPU, and is approximately 80~MB, making it compatible with the vLLM server
that holds the GPU in exclusive process mode. The embedder is run on CPU to
avoid a device ownership conflict with vLLM.

At inference time, the user query is embedded and the top-5 most similar
functions are retrieved as candidates ($k=5$). The value $k=5$ is selected to
achieve near-complete recall while keeping the candidate set small enough for
the model to disambiguate: a smaller $k$ risks excluding the correct function,
while a larger $k$ increases the number of semantically similar alternatives
the model must distinguish. The model selects among the retrieved candidates
using \texttt{guided\_choice} restricted to the five retrieved function names,
then extracts arguments using \texttt{guided\_json} as in CD+Q. All
other settings are inherited from CD+Q.

\section{LoRA Fine-Tuning}

Two fine-tuning configurations are evaluated: FT-only, which applies
the trained model without constrained decoding, and CD+FT, which
combines the trained model with the constrained decoding pipeline from
CD. The split between the two variants allows the independent
contributions of fine-tuning and constrained decoding to be measured
separately.

The base model is \texttt{Qwen/Qwen2.5-7B-Instruct} in bfloat16. A LoRA
adapter is trained for 2~epochs with rank~16 on the Salesforce
\texttt{xlam-function-calling-60k} dataset~\cite{zhang2024xlam}, using
54{,}000 training examples and 6{,}000 validation examples, with the
HuggingFace PEFT library and TRL SFTTrainer~\cite{hu2022lora}.

Rank~16 is selected as a mid-range value for instruction-following adaptation:
it provides sufficient capacity for the model to learn format and value
conventions from the training data without overfitting. Two epochs are selected
because the dataset is large enough that a single pass provides substantial
exposure to the training distribution; a third epoch produced no further
improvement in validation loss in preliminary runs.

After training, the adapter weights are merged into the base model by linear
interpolation, producing a single merged checkpoint in bfloat16. The merged
model is used for all evaluation runs, removing adapter inference overhead and
ensuring that the evaluation infrastructure is identical to that used for
Base and CD.

\section{Schema-Enriched Constrained Decoding}

Config~CD+schema extends CD by enriching the argument extraction prompt
with the full parameter schema detail available in the BFCL function
definitions. In all preceding configurations, the Stage~2
\texttt{guided\_json} prompt carries only parameter names and types.
Config~CD+schema additionally injects, for each parameter, its natural-language
description, default value (if any), and enumeration constraints (if any),
drawn directly from the BFCL function schema.

The structural constraint is unchanged. \texttt{guided\_json} continues
to enforce JSON object validity and type compliance at each generation
step via the same finite-state machine as CD. The schema enrichment
acts on the prompt only: it provides the model with richer semantic
context for value selection without altering the token-level mask.
The two mechanisms are composable and independent.

Figure~\ref{fig:template-vs-enriched} shows the difference on a single
test case (\texttt{simple\_python\_34}). The template prompt lists each
parameter only as \texttt{name (type)}; the enriched prompt adds a
\texttt{Parameters:} block carrying the natural-language description,
the allowed values, and the default for each parameter. Here the user
asks to ``prefer exploring nature'' while the \texttt{exploration\_type}
parameter is an enumeration whose members
(\texttt{nature}, \texttt{urban}, \texttt{history}, \texttt{culture})
appear only in the enriched prompt: under the template prompt the model
must guess the exact enum spelling, whereas the enriched prompt supplies it.

\begin{figure}[ht]
\textbf{(a) CD --- template prompt (\texttt{name (type)} listing only)}
\begin{lstlisting}[style=promptbox]
Function: travel_itinerary_generator(destination: string, days: integer, daily_budget: integer, exploration_type: string) -> string
Description: Generate a travel itinerary based on specific destination, duration and daily budget, with preferred exploration type.
User request: Create an itinerary for a 7 days trip to Tokyo with daily budgets not exceeding $100 and prefer exploring nature.
Extract the argument values as a JSON object with keys: destination (string), days (integer), daily_budget (integer), exploration_type (string).
JSON:
\end{lstlisting}

\vspace{4pt}
\textbf{(b) CD+schema --- enriched prompt (per-parameter schema block added)}
\begin{lstlisting}[style=promptbox]
Function: travel_itinerary_generator(destination: string, days: integer, daily_budget: integer, exploration_type: string) -> string
Description: Generate a travel itinerary based on specific destination, duration and daily budget, with preferred exploration type.
Parameters:
- destination (string, required): Destination city of the trip.
- days (integer, required): Number of days for the trip.
- daily_budget (integer, required): The maximum daily budget for the trip.
- exploration_type (string, optional): The preferred exploration type. Allowed values: "nature", "urban", "history", "culture". Default: "urban".
User request: Create an itinerary for a 7 days trip to Tokyo with daily budgets not exceeding $100 and prefer exploring nature.
Extract the argument values as a JSON object with keys: destination (string), days (integer), daily_budget (integer), exploration_type (string).
JSON:
\end{lstlisting}
\caption{Template versus schema-enriched Stage~2 argument-extraction prompt
for the same test case (\texttt{simple\_python\_34}). \textbf{(a)} The CD
template prompt carries only the reduced \texttt{name (type)} parameter
listing. \textbf{(b)} Config~CD+schema inserts a \texttt{Parameters:} block
with each parameter's description, enumerated values, and default, drawn
directly from the BFCL function schema. The \texttt{guided\_json} constraint
is identical in both cases; only the prompt gains information. The enumerated
values for \texttt{exploration\_type}, which the template prompt omits, are
exactly what the model needs to map ``prefer exploring nature'' onto the
literal \texttt{"nature"}.}
\label{fig:template-vs-enriched}
\end{figure}

Config~CD+schema is evaluated on the full 400-case
\texttt{simple\_python} test set across all four Qwen~2.5 sizes for the
size-scaling analysis. The primary paired significance test is reported at
7B, where a paired McNemar test on stored per-case predictions compares
CD+schema against the stored CD run (290/400) used as the paired baseline.
The size sweep shows whether the schema-enrichment effect persists across
capacity, while the paired 7B comparison isolates the contribution of schema
information to the residual failure rate after constrained decoding.
