\chapter{Discussion} \label{ch:discussion}

This chapter interprets the experimental results in terms of the thesis
questions. The central distinction is between structural failures, which can
be removed by constrained decoding, and semantic value errors, which require
better schema information, training data, or routing. The chapter first
analyses this distinction, then discusses cost-benefit trade-offs, limitations,
practical deployment implications, and the relation to prior work.

\section{Impact of Constrained Decoding vs.\ Model Scale}

The gap between Base and CD is almost entirely a format compliance gap. Under
the strict whole-completion parser this gap is 71.25 percentage points; under
the lenient first-object parser used in the isolation arm, the 7B same-parser
accuracy gain is 10.75~pp. Of the 394 strict-parse failures in Base, virtually
all are structural: the model produces conversational text, multiple JSON
objects, or empty output rather than a single valid function call. The six
correct outputs in Base are semantically accurate but happen to be formatted
correctly by chance. The model is not unable to reason about the queries; it
cannot express its answers in the required structured format without guidance.

This distinction has a direct consequence for interpreting the frontier
comparison. CD achieves 72.75\% on BFCL~v4 \texttt{simple\_python},
6.00~pp below the weakest frontier entry on the same category (GPT-5-mini,
78.75\%) and 18.25--25.00~pp below flagship frontier models
(91.00\%--97.75\%)~\cite{bfcl2024}.\footnote{Frontier model scores are the Python Simple AST sub-category of the BFCL~v4 leaderboard, accessed 2026-06-04 and re-verified 2026-06-10; see the comparability note in Section~\ref{sec:frontier-comparison}.} Constrained decoding enforces valid JSON and valid function
names by construction, which is precisely what frontier models' native
tool-calling APIs also do. Once format compliance is equalized, a 7B model
and a frontier model are distinguished only by argument value accuracy.
There the constrained pipeline trails the flagship models by 14~pp or
more, while the same model under its native template (B-template, 96.00\%)
is competitive with them: the deficit belongs to the model-agnostic
pipeline, not to the model.

The comparison is specific to the \texttt{simple\_python} category. Each test
case provides exactly one candidate function, eliminating function selection
as a source of error. With one candidate and \texttt{guided\_choice},
selection accuracy is 100\% regardless of model size. The remaining 27.25\%
failure rate is entirely in argument values: wrong optional parameter defaults,
numeric precision errors, and string format mismatches. Frontier models make
the same class of errors at a lower rate; the gap of 25.00 percentage points
to the best frontier model (Claude Sonnet~4.5, 97.75\%) reflects a higher
frequency of correct value conventions under prompts that carry the full
schema detail, not a qualitatively different failure mode.

The implication for RQ1 is that the raw gap of approximately
96 percentage points between the model-agnostic baseline (7B Base at 1.5\%)
and the best frontier model (Claude Sonnet~4.5 at 97.75\%)
collapses to approximately 25 percentage points with constrained decoding
alone, on this benchmark category. Two readings of this gap must be kept
apart. The 1.5\% baseline characterises a generic integration without the
model's native tool-calling template; the B-template control measures the
same model at 96.00\% with native prompting and free generation
(Chapter~\ref{ch:methodology}), so the 96~pp figure measures the cost of
naive integration, not a capability deficit of the model. What the
constrained decoding result adds is that format compliance can be restored
structurally, with guarantees independent of any model-specific template.
Scale is not the bottleneck for format compliance. The residual gap is a
semantic problem, but the B-template control constrains its
interpretation: given its native template and the full schemas, the model
resolves most of these cases. Config~CD+schema isolates the information-loss
component directly: enriching the constrained prompt with parameter
descriptions, enumeration values, and defaults raises accuracy from
72.75\% to 89.00\% ($+$16.25~pp, McNemar $p < 0.00001$;
Section~\ref{sec:cd-schema}), confirming that the majority of the residual
gap after plain CD is attributable to missing schema detail rather than to
the model's learned value priors. The remaining 7.00~pp gap between
CD+schema (89.00\%) and B-template (96.00\%) reflects the prompt-format
advantage of the native template and residual value-convention scorer
artifacts; it is not dominated by information loss.

\subsection{Effect of Model Scale on CD Performance}

The format compliance failure that accounts for 98.5\% of Base errors
is not resolved by scale.
Without constrained decoding, accuracy across the 0.5B, 1.5B, 3B,
and 7B Qwen~2.5 models ranges from 1.50\% to 4.75\%, despite a
14-fold difference in parameter count.
The six correct outputs in 7B Base and the fourteen correct outputs
in 0.5B Base are produced by chance; the underlying failure mode,
malformed or conversational output rather than a single valid JSON
object, is present at every scale.

With constrained decoding applied, the accuracy curve follows a
consistent scaling trend: 51.50\% at 0.5B, 62.25\% at 1.5B, 64.75\%
at 3B, and 72.75\% at 7B.
The steepest gain occurs between 0.5B and 1.5B (+10.75~pp); the
1.5B-to-3B step is flatter (+2.5~pp); the 3B-to-7B step recovers
to +8.0~pp.
Constrained decoding shifts the entire curve upward without altering
its shape, consistent with the view that format compliance and
semantic accuracy are separable and that the scaling relationship
governs the latter.

The AWQ quantization penalty is inversely related to model size.
The accuracy loss relative to the bfloat16 baseline is 3.75~pp at 0.5B,
1.75~pp at 1.5B, 0.25~pp at 3B, and 0.50~pp at 7B.
At 3B and above, the INT4 approximation falls within run-to-run
variance.
At 0.5B, the 3.75~pp loss reflects the reduced parameter redundancy
available to absorb quantization error, consistent with the
sensitivity of AWQ calibration on small weight
matrices~\cite{lin2024awq}.
Quantization is therefore safe for deployment above 1.5B and
introduces a meaningful but bounded cost at 0.5B.

Few-shot prompting (Config~PE) reverses direction at 7B.
At 0.5B, 1.5B, and 3B, the three in-context examples improve
accuracy by 2.0 to 2.5~pp over the CD baseline, suggesting the
examples compensate for weaker internal argument-value
representations at smaller scales.
At 7B, the same examples reduce accuracy by 2.5~pp.
The approximately 300 additional context tokens introduced by the
examples appear to compete with the test case for attention at 7B,
an effect consistent with the prompt-length sensitivity reported
for large-scale in-context learners~\cite{brown2020language}.
The practical implication is that the best no-training configuration
is size-dependent: Config~CD alone at 7B, and Config~PE at
sizes up to 3B.

\section{Cost-Benefit Analysis of Each Method}

Constrained decoding yields the highest structural return of any technique
evaluated. It guarantees parseable outputs with no training, no extra
parameters, and no additional generated tokens. The accuracy gain is 71.25
percentage points relative to the strict generic floor and 10.75~pp relative to
the lenient 7B parser baseline. The per-step cost of guided decoding is not
zero: schema compilation and logit masking add overhead that grows with schema
complexity, although the finite-state machine approach keeps the per-token cost
low when schemas are compiled ahead of time~\cite{willard2023efficient}.
Measured on 100 sequential \texttt{simple\_python} cases
(Section~\ref{sec:latency}), constrained decoding (0.878~s mean per call) is
8$\times$ faster than the unconstrained baseline (6.995~s mean): the
unconstrained model appends prose, markdown, and additional JSON objects,
generating far more tokens; the constraint terminates generation at the closing
brace. Constrained decoding is therefore not a latency tax but a latency gain
at 7B. The structural gain comes from logit masking at each decoding step,
which eliminates the malformed output that accounts for 98.5\% of failures in
Base~\cite{willard2023efficient, kwon2023vllm}.

Schema-enriched constrained decoding extends the plain CD result by supplying
the parameter descriptions, type constraints, and enumerated values that the
standard pipeline omits from its prompts. Config~CD+schema achieves 89.00\%
($+16.25$~pp over plain CD, McNemar $p < 0.00001$;
Section~\ref{sec:cd-schema}), the highest accuracy of any configuration in
the evaluation and the only result to enter the lower frontier range on BFCL~v4 Python Simple AST without
modifying model weights. No training, no additional generated tokens, and no
latency overhead are introduced: the measured per-call latency for CD+schema
(0.704~s mean) falls within the same range as plain CD and does not exceed
the unconstrained baseline. The implementation cost is maintaining enriched
prompt templates that include parameter descriptions, type annotations, and
enumerated defaults for each tool. The cost-benefit ratio is the most
favorable in the evaluation, and the gain scales with schema richness: tools
with detailed descriptions and enumerated defaults expose more of the missing
vocabulary that schema enrichment recovers.

AWQ INT4 quantization adds a further infrastructure benefit at near-zero
accuracy cost~\cite{lin2024awq}. The $-0.5$ percentage point change (two
test cases) falls within run-to-run variance. Memory is reduced by 63.5\%
(14.25~GiB to 5.20~GiB) and model loading time by 83.5\% (212~s to 35~s).
The cost-benefit ratio is strongly positive: substantial deployment benefits
at negligible accuracy cost.

Few-shot prompting (PE) and chain-of-thought reasoning
(CD+Q+ITC) both produce negative outcomes~\cite{brown2020language,
wei2022chain}. PE reduces accuracy by 2.5 percentage points with a
context overhead of approximately 300 tokens per query. CD+Q+ITC
reduces accuracy by 6.75 percentage points while increasing per-case
inference time proportionally to the additional tokens generated by the
reasoning trace (Chapter~\ref{ch:results} details the per-case flip
analysis and the three mechanisms behind the CoT regression). Both
techniques share the same root cause: the residual 27\% failure rate
after constrained decoding is a mismatch between the values the
pipeline elicits and the benchmark's ground truth conventions. The
B-template control indicates that a substantial part of this residual
stems from schema information missing from the pipeline's prompts:
neither three in-context examples nor an explicit reasoning step
substitutes for the absent parameter descriptions, and the reasoning
step actively moves values away from the benchmark's conventions. Two
independent negative results with the same directional outcome
establish that prompt-level additions within the pipeline cannot
address this failure class; supplying the missing schema detail itself,
as the native template does, is a different and evidently more
effective intervention.

This finding aligns with the scope conditions reported by Wei et
al.~\cite{wei2022chain}: chain-of-thought prompting improves performance
on tasks requiring multi-step arithmetic or symbolic decomposition, not
on tasks where the bottleneck is value recall. Argument extraction is
not a reasoning problem in that sense; an explicit reasoning step
directs compute at the wrong subproblem~\cite{snell2024scaling}.

Retrieval-augmented generation produces the largest accuracy drop: $-24.5$
percentage points relative to CD+Q, despite a retrieval recall@5 of
97.2\%~\cite{lewis2020retrieval}. The bottleneck is not retrieval quality
but candidate disambiguation. When multiple semantically similar functions
are retrieved, the 7B model cannot reliably identify the one matching the
user's intent. The infrastructure cost of RAG is substantial (an offline
FAISS index and an embedder at query time), making its negative accuracy
outcome particularly unfavorable.

The disambiguation failure identified here is avoided by an alternative
design in Gorilla~\cite{patil2023gorilla}, which incorporates retrieval at
training time rather than inference time. By generating training examples from
retrieved API documentation, Gorilla teaches the model to use retrieved
content as a grounding signal without exposing it to multiple similar
candidates at query time. This sidesteps the inference-time disambiguation
problem entirely, at the cost of coupling the model to a fixed API catalog.
The configuration evaluated in this thesis uses semantic similarity retrieval
at inference time with no training signal on candidate disambiguation, which
may explain the low selection accuracy despite high retrieval recall@5
(97.2\%). The 66\% of all test cases whose output is identical to the non-RAG
prediction (Fig.~\ref{fig:rag-breakdown}) confirms that retrieval adds
overhead without changing the model's output for the majority of cases. A design that combined
inference-time retrieval with a fine-tuning signal on tool selection from
retrieved candidates would be a natural next step.

LoRA fine-tuning on the \texttt{xlam-function-calling-60k}
dataset~\cite{zhang2024xlam, hu2022lora} produces a regression of 3
percentage points under CD+FT relative to CD. The regression
is attributed to a format mismatch between the training distribution, which
encodes assistant turns as Python call syntax, and the evaluation pipeline,
which expects a JSON argument object. The technique is not ruled out by this
result. The format mismatch hypothesis is confirmed by CD+FT-aligned,
which retrains on the same xlam content reformatted to match the inference
pipeline's JSON output convention exactly. CD+FT-aligned achieves
76.75\% (307/400), a gain of 7.00 percentage points over CD+FT v1 and
4.00 percentage points above the CD baseline, making it the best
training-based result in the evaluation. Under the corrected lenient parser,
plain fine-tuning does not raise unguided
format compliance above Base: FT-only reaches 53.00\%, below the no-guided
Base (62.00\%), and FT-aligned-ng is lower still at 41.00\%. The 35.75
percentage point gap between FT-aligned-ng (41.00\%) and CD+FT-aligned
(76.75\%) shows that constrained decoding remains the major contributor even
after fine-tuning. The same dependence holds across families and sharpens
away from the 7B setting: without guided decoding the format-aligned adapters
fall to 0.00\%--3.75\% (Phi-4-mini reaches 0.00\%), well below the 39.00\%--67.50\%
retained by the matched instruct checkpoints, so format alignment specialises
the model to the constrained-decoding protocol rather than producing a
stronger standalone generator.

Taken together, the results support a clear ordering by cost-benefit ratio:
schema-enriched constrained decoding (CD+schema, 89.00\%) delivers the highest
accuracy at zero training cost; plain constrained decoding provides the core
structural gain with no overhead; quantization adds substantial memory and
loading benefits at near-zero accuracy cost; and format-aligned fine-tuning
offers further marginal gains at the highest implementation cost. Prompting
and retrieval interventions produce negative outcomes on this failure class.

\section{Limitations}

The results are based on the \texttt{simple\_python} category of BFCL~v4,
which is the least demanding of the four BFCL categories~\cite{bfcl2024}.
Each test case consists of a single user turn, a single candidate function,
and a single expected call. The constrained decoding advantage depends on
oracle function provision: \texttt{guided\_choice} guarantees correct
function selection only when the correct function is the sole candidate.
Generalisation to harder categories is partial: CD+FT-aligned reaches
70.5\% on \texttt{multiple} at 7B, and the \texttt{parallel} category
exposes a limitation in the output schema rather than in model
capability. Under the initial single-extraction design the
\texttt{parallel} score is 0.0\% across all configurations and sizes,
and this is not a soft accuracy limitation that more training or a
larger model would gradually close; it follows from the structure of
the parallel selection stage described in Chapter~\ref{ch:methodology}.
That stage emits a set of distinct function names and runs one
deterministic argument extraction per selected name, so the pipeline can
produce at most one call per distinct function. In the \texttt{parallel}
category, where every required call targets the same single candidate
function, the pipeline emits exactly one call by construction; in
\texttt{parallel\_multiple}, where the required calls target distinct
functions, the same pipeline produces up to one call per function,
which is why multi-call responses are observed there
(Section~\ref{sec:bfcl-extended}). Supporting repeated calls to the
same function requires an output schema in which a JSON array holds
complete call objects with their own argument bindings, or a multi-step
execution loop where each iteration emits one call.

The first of these remedies was implemented and evaluated. Replacing the
single-extraction stage with a joint multi-call schema, in which the model
generates an array of complete call objects in one guided pass, lifts
\texttt{parallel} from a structural 0.0\% to between 9.5\% at 0.5B and 79.0\%
at 7B (Section~\ref{sec:bfcl-extended}). This confirms that the constraint
was the schema rather than model scale: once repeated calls are permitted,
accuracy on the category depends on size in the same way as the other
categories. The result therefore identifies a boundary of the
single-extraction architecture and the effect of moving past it, rather than
a permanent ceiling on the task; same-function parallel calling at the
smallest sizes remains hard, so the category is not solved outright.

The size sweep and most ablation configurations use a single model
family (Qwen~2.5) across four sizes. Constrained decoding is the
exception: it is also evaluated on four other families
(Section~\ref{sec:cross-family-cd}), where its structural benefit holds,
so that finding has external support. The remaining directional findings,
such as the ordering of techniques by accuracy, the size-dependence of
the PE effect, and the quantization penalty profile, are observed
consistently within Qwen~2.5 only. Whether they transfer to other SLM
families with different training regimes (e.g.\ Llama~3.2, Phi-4) is not
established and should not be assumed.

The cascade architecture is the motivating deployment context for this
evaluation, but no direct cascade experiment is reported. The benefit
fractions ($p \approx 0.72$--$0.77$) cited for each configuration are
extrapolated from single-model BFCL accuracy; actual cascade economics
depend on routing accuracy, latency characteristics, and the query
distribution of the production workload, none of which were measured.

The BFCL ground truth encodes specific value conventions that are often
arbitrary from a deployment perspective. A model that produces
\texttt{"gasoline"} where the ground truth expects \texttt{"gas"} is scored
as incorrect despite both being valid in a real API call. The 27\% residual
failure rate after constrained decoding may overestimate the practical error
rate for production use, where value conventions are defined by the API
schema rather than by a benchmarking dataset.

Multi-step performance ($\tau$-bench: 4.35\% pass rate) is evaluated for
CD at 7B and for CD and CD+FT-aligned across the size sweep, but not for
the full configuration set. The $\tau$-bench result establishes that
single-call accuracy does not predict multi-step agentic performance; the
remaining no-training configurations were not evaluated on $\tau$-bench
because the bottleneck identified is multi-turn planning rather than
argument formatting.

\subsection{Threats to Validity}
\label{sec:threats-validity}

Several factors limit the generalisability of the findings.

\textbf{Baseline construction and schema information loss.}
The Base configuration is a minimal model-agnostic integration that
does not use Qwen~2.5's native tool-calling prompt template. The
B-template control measures the same model at 96.00\% (384/400) under
native prompting with free generation
(Table~\ref{tab:baseline-results}), so the headline gap to
frontier models in RQ1 quantifies the cost of naive integration
rather than an intrinsic capability deficit. The information-loss
component of this gap is now quantified directly by Config~CD+schema
(Section~\ref{sec:cd-schema}): enriching the constrained prompt
with the full schema detail raises accuracy from 72.75\% to 89.00\%
($+$16.25~pp, McNemar $p < 0.00001$), demonstrating that most of the
residual after plain CD is attributable to the absent descriptions,
enumerations, and defaults rather than to the model's learned value
priors. The remaining 7.00~pp gap between CD+schema and B-template
is attributable to the native template's prompt-format advantage and
residual value-convention scorer artifacts.

\textbf{$\tau$-bench user simulator.}
The user simulator is the agent model itself (Qwen~2.5 7B), whereas
the reference $\tau$-bench setup uses a frontier-model simulator.
Simulator errors (misstated or omitted task details) deflate pass
rates independently of agent capability, so the absolute pass rate
conflates agent and simulator failures, and comparisons to published
pass rates under stronger simulators are not like-for-like.
Three independent runs at identical agent settings (temperature~0.0,
seed~42) produced pass rates of 3.48\%, 4.35\%, and 5.22\%
(4--6 tasks out of 115). Because the agent is fully deterministic
across runs, the 1.74~pp spread is attributable entirely to stochastic
simulator turns, bounding the observed simulator-noise contribution at
$\pm 0.87$~pp (one task at $n{=}115$). The qualitative conclusion, that
single-call BFCL gains do not transfer to multi-step agentic
completion, is robust to this effect: even at the upper bound (5.22\%)
the gap to the 72.75\% BFCL CD result exceeds 67~pp, and the
conclusion holds across all evaluated model sizes
(Table~\ref{tab:tau-bench-size}).

\textbf{Single model family.}
The size sweep and the quantized, fine-tuned, and agentic configurations
use the Qwen~2.5 Instruct family across four sizes. Constrained decoding
is additionally evaluated on four other families
(Section~\ref{sec:cross-family-cd}), where its benefit is family-robust,
and the GitHub MCP and format-aligned analyses also span families
(Sections~\ref{sec:github-mcp} and~\ref{sec:format-aligned}). The
remaining findings (the ordering of techniques by accuracy, the
size-dependence of the few-shot effect, and the quantization penalty
profile) are observed only within Qwen~2.5 and may not transfer to other
SLM families with different pre-training regimes or instruction-tuning
data (e.g.\ Llama~3.2, Phi-4).

\textbf{BFCL \texttt{simple\_python} reliance.}
The primary evaluation uses the least demanding BFCL category:
single-turn, single-candidate queries with no function-selection
ambiguity. Results on harder categories are partial (Section~\ref{sec:bfcl-extended}),
and the constrained decoding advantage depends on oracle function
provision. Generalization to multi-turn or high-ambiguity settings
cannot be assumed.

\textbf{Frontier scores not re-evaluated locally.}
Frontier model scores are taken from the official BFCL leaderboard
rather than re-run under the same evaluation harness. Minor
methodology differences (prompt formatting, API tool-call handling)
may affect comparability, so the frontier comparisons indicate parity
rather than a definitive ranking.

\textbf{Accuracy as routing proxy.}
BFCL accuracy is used to proxy the cascade traffic fraction $p$.
After constrained decoding removes structural failures, every SLM
output is a valid JSON object naming a known function; structural
routing signals are no longer available. The reported $p$ values are
upper-bound estimates that assume an ideal semantic router.

\textbf{Single-run protocol and run-to-run spread.}
BFCL results use temperature~0; repeated runs within a fixed software
environment produce identical outputs (six-run confirmation for CD+Q),
but re-runs spread over the project period varied by one to two cases
out of 400, an uncertainty of roughly $\pm$0.5~pp on any single
accuracy figure. Differences of this magnitude between configurations
(for example, the 0.5~pp CD-to-CD+Q delta) are therefore not
individually meaningful, and the paired tests in
Chapter~\ref{ch:results} are interpreted against this spread. For
$\tau$-bench, three runs are completed at 7B but only single runs at
smaller sizes, where one-task differences fall within the noise band.

\textbf{Benchmark convention mismatch.}
The 27\% residual failure rate after constrained decoding may
overestimate practical error rates. BFCL ground truth encodes
specific value conventions (e.g.\ \texttt{"gas"} not
\texttt{"gasoline"}, decimal fractions not percentages) that a
production API schema would not impose. The residual failures
represent convention mismatches, not functional errors, in many cases.

\textbf{Few-shot example design.}
The three PE examples (Chapter~\ref{ch:methodology}) were authored to target
failure classes identified by inspecting the test-split errors, so the design
of the few-shot technique is informed by the evaluation set. This biases in
PE's favour. The weak and size-dependent PE effect (at best $+2.5$~pp, and
$-2.5$~pp at 7B) is therefore a conservative estimate rather than an inflated
one: the technique fails to help even when its examples are tuned to the
observed errors. A held-out development split would be the correct design for
a clean, unbiased measurement.

\section{Practical Implications} \label{sec:practical-implications}

The cascade architecture motivating this work assumes that an SLM handles
most tool-calling requests locally and escalates only the unreliable cases to
a frontier model API. The results make this architecture possible to cost, but
only as a projection. No cascade router is implemented here. The estimates
below presume a semantic router that can identify likely-wrong SLM outputs;
after constrained decoding, structural routing signals are unavailable because
every output is valid JSON naming a known function. The reported $p$ values
are therefore upper bounds.

As a first-order projection, suppose frontier API calls cost $c$ per query and
the local SLM tier has already been deployed, so its marginal cost is treated
as small relative to the API call. Under that simplifying assumption, a cascade
in which the SLM correctly handles a fraction $p$ of requests saves about
$p \cdot c$ per query relative to routing all traffic to the frontier model.
CD+Q achieves $p \approx 0.72$, CD+Q+FT-aligned achieves $p \approx 0.74$, and
CD+FT-aligned achieves $p \approx 0.77$, all on a single consumer GPU\@. These
are upper-bound traffic fractions, not measured deployment savings, because
the router, utilization, batching, energy cost, maintenance cost, and latency
service level are not evaluated. The cascade criterion of RQ3
(Chapter~\ref{ch:introduction}) adopts $p \geq 0.70$ as its reference point,
which all four constrained configurations at 7B exceed under this proxy.

The cost-benefit framing above costs the constrained local tier against a
frontier API. For a single-model deployment a stronger local option is
available: the same 7B model under its native tool-calling template
(B-template) reaches 96.00\% on \texttt{simple\_python}
(Table~\ref{tab:baseline-results}), above every constrained configuration
on the same hardware. For a single frozen Qwen deployment, the native template
is therefore the better local baseline. The constrained pipeline is valuable
when the goal is model-agnostic structural guarantees across heterogeneous
models and schemas: output validity is guaranteed independently of any model's
template, a uniform integration is retained, and the router receives
guaranteed-parseable output. The $p$ values reported in this section thus apply
to the general multi-model case; for a single frozen model the native template
supersedes them.

These values are upper-bound estimates: after constrained decoding is
applied, every SLM output is structurally valid, removing structural
routing signals. Achieving $p$ close to the measured accuracy requires a
semantic routing mechanism that identifies likely semantic failures before
or after inference; such a mechanism is not evaluated in this thesis.
The marginal benefit of each additional technique is therefore expressible
directly in terms of API cost reduction: a 2.0~pp gain from adding
quantized fine-tuning on top of CD+Q eliminates 2.0\% of remaining
frontier API calls per unit time at high query volume.

Whether this gain justifies the training investment depends on query
volume. At low throughput, the API cost reduction does not accumulate
quickly enough to recover the fine-tuning compute cost; the no-training
CD+Q configuration is preferable. At high throughput, the per-query
saving compounds and the break-even point moves earlier in the
deployment lifecycle. The break-even query volume is given by
$N = C_{\text{train}} / (\Delta p \cdot c_{\text{frontier}})$,
where $C_{\text{train}}$ is the one-time training cost, $\Delta p$ is
the accuracy gain, and $c_{\text{frontier}}$ is the frontier API cost
per call. For CD+Q+FT-aligned versus CD+Q, $\Delta p \approx 0.02$. The two cost terms can
be grounded in current list prices rather than left illustrative. A frontier
tool call to GPT-4o is billed at \$2.50 per million input tokens and \$10.00 per
million output tokens.\footnote{\url{https://openai.com/api/pricing/}, accessed
2026-06-28.} A representative \texttt{simple\_python}-style call consisting of
roughly 2{,}000 input tokens (system instruction, tool schemas, and user query)
and 150 output tokens (a single JSON call) therefore costs
$c_{\text{frontier}} \approx \$0.0065$. The one-time training cost is the
approximately 4~GPU-hours of the LoRA run priced at a current A100~80\,GB cloud
rate of about \$1.50 per GPU-hour,\footnote{Representative on-demand rate for a
single A100~80\,GB; community-cloud providers quote \$1.19--\$1.49 per GPU-hour
(\url{https://www.runpod.io/pricing}, accessed 2026-06-28).} giving
$C_{\text{train}} \approx \$6$. These figures place the break-even point at
$N = C_{\text{train}} / (\Delta p \cdot c_{\text{frontier}}) \approx 46{,}000$
queries. The number is not a deployment cost estimate: it omits local serving
utilization, queueing, energy, monitoring, router cost, and engineering
maintenance. It is also highly sensitive to the escalation target. If the
frontier tier is a cheaper model such as GPT-4o-mini (\$0.15 and \$0.60 per
million input and output tokens), the same call costs only
$c_{\text{frontier}} \approx \$0.0004$ and the break-even rises above 750{,}000
queries; the trained tier pays for itself only when the model it displaces is an
expensive one. At lower frontier costs or smaller query volumes the no-training
configuration remains preferable; at higher volumes against a frontier-priced
target the investment recovers quickly. The 2.0--5.0~pp accuracy range separating the local-tier
configurations represents a concrete trade-off space.

The residual failure classes common to all configurations (wrong optional
parameter defaults, numeric precision mismatches, and string format
conventions) define the queries that will be escalated regardless of which
configuration is deployed. Structural output validation, which would be
the natural first routing signal, is ineffective after constrained decoding
is applied: every output is valid JSON naming a known function. The
relevant routing signals are semantic: query-level classification of
likely hard cases before inference, or token-probability confidence scoring
after inference. Both are tractable extensions of the evaluated pipeline
and represent the primary mechanisms for narrowing the escalation boundary
beyond what the optimization techniques evaluated here can achieve.

Concretely, such a router would sit between the constrained pipeline and the
frontier API and operate on two signals. The first is a pre-inference query
classifier: a lightweight model, for example a sentence embedding of the user
query and the candidate tool schema followed by a logistic regression, trained
on the per-case correct/incorrect labels already collected in this evaluation
to predict the probability that the SLM will produce a semantically correct
call, and escalating queries above a risk threshold without running the SLM at
all. The second is a post-inference confidence signal derived from the decoder
itself. Although constrained decoding guarantees that every output is valid
JSON, the token-level log-probabilities under the guided mask are not uniform:
the mean log-probability over the extracted argument tokens, or the entropy at
the masked argument positions, provides a calibratable score that a threshold
can convert into a route decision. A complementary post-inference signal is
self-consistency, in which the argument extraction is sampled more than once and
disagreement among the samples triggers escalation. Because the residual failure
classes are systematic (optional-parameter defaults, numeric precision, and
string-format conventions) rather than random, a classifier trained on these
labelled cases is the more promising of the two: it targets a structured rather
than a diffuse error distribution. The threshold in either case sets the
operating point on a precision-recall trade-off between API cost saved and
incorrect calls served locally, and its calibration is the design parameter that
determines how close the realised $p$ comes to the upper bounds reported above.

Fine-tuning alignment is the binding constraint on the achievable accuracy
ceiling. The xlam experiment demonstrates that a general-purpose
function-calling corpus with a different output convention actively degrades
performance under constrained decoding: the model learns value priors from
the training distribution that conflict with the evaluation pipeline's
format. This means that the fine-tuning gain is not portable across
pipelines. A production fine-tuning run must generate training examples
from the target deployment's own tool schemas; otherwise the format
mismatch replaces the original semantic failure class with a new one.

\section{Relation to Prior Work}

The results reported in this thesis intersect four bodies of prior work:
constrained generation, training-based tool adaptation, retrieval-augmented
generation for tool selection, and multi-step agentic evaluation.

\paragraph{Constrained generation.}
Willard and Louf formalised constrained generation using finite-state
machines compiled from JSON schemas, establishing that format compliance can
be guaranteed at each decoding step with low per-step
overhead~\cite{willard2023efficient}. The Base-to-CD comparison confirms this
guarantee in practice: every failure involving malformed or conversational
output under the strict Base parse is eliminated, and the residual 27.25\%
failure rate is attributable entirely to argument value errors. This validates
the scope claim in Willard and Louf: constrained decoding solves the format
compliance problem completely but cannot address semantic accuracy. The
strict-parser accuracy recovery is 71.25~pp at 7B, but the same-parser gain over
the lenient first-object baseline is 10.75~pp. The value of constrained decoding
here is therefore the structural guarantee first and the accuracy margin second;
models or parsers that already recover valid JSON would show a smaller measured
accuracy benefit.
The negative chain-of-thought interaction observed in
Section~\ref{sec:no-training} also connects to the study of Tam et
al.~\cite{tam2024letmespeak}, which reports that format restrictions can
degrade reasoning performance: the present results show the complementary
direction, namely that adding explicit reasoning under a format constraint
degrades value accuracy when the bottleneck is convention recall rather
than reasoning. Newer structured generation engines such as
XGrammar~\cite{dong2024xgrammar} reduce the per-step overhead of
grammar-constrained decoding but do not change the accuracy trade-offs
analysed here.

\paragraph{Training-based tool adaptation.}
Gorilla demonstrated that fine-tuning a LLaMA model on retrieved API
documentation can outperform GPT-4 on API-calling tasks, establishing that
training data relevance can compensate for parameter count
disadvantage~\cite{patil2023gorilla}. The key mechanism is alignment between
the model's training context and its inference context. This thesis extends
that insight from API source selection to output format alignment. Two LoRA
adapters trained on the same xLAM dataset~\cite{zhang2024xlam} with identical hyperparameters
differ by 7.0~pp on BFCL solely because their training output formats
conflict or match the inference pipeline's JSON target.
The format alignment effect is of comparable magnitude to the improvements
Gorilla reports from retrieval-augmented training, suggesting that output
format compatibility is as important a dimension of training data design as
content relevance.

A broader survey of small language models for agentic systems identifies
Qwen~2.5 7B as one of the strongest open-weight SLMs for tool use and notes that
systematic combinations of constrained decoding, quantization, and fine-tuning
remain underexplored~\cite{sharma2025slmagents}. The cumulative evaluation
design adopted in this thesis directly addresses that gap, providing
comparative measurements across five technique combinations applied to a
single model family over four model sizes.

\paragraph{Retrieval-augmented generation for tool selection.}
Lewis et al.\ showed that retrieval-augmented generation improves performance
on knowledge-intensive tasks by supplying relevant context at inference
time~\cite{lewis2020retrieval}. The RAG evaluation here reveals a limitation
of this mechanism for tool selection: retrieval recall@5 of 97.2\% produces a
$-24.5$~pp accuracy drop relative to CD+Q rather than an accuracy gain, because the bottleneck
shifts from retrieval quality to candidate disambiguation. Retrieved candidates
are semantically similar by construction, and the model must distinguish
between them on fine-grained semantic grounds that single-call training does
not develop. This failure mode is consistent with ToolLLM's observation that
model performance degrades as the number of candidate APIs
grows~\cite{qin2024toolllm}. Gorilla's training-time retrieval approach avoids
the problem by teaching the model to handle similar candidates during training;
inference-time retrieval without a corresponding disambiguation signal does not
produce the same effect.

\paragraph{Multi-step agentic evaluation.}
The 68.4~pp gap between BFCL single-call accuracy (72.75\%) and $\tau$-bench
pass rate (4.35\%) reveals a discontinuity between structured output generation
and multi-step reasoning. The ReAct framework and related agentic approaches
report a similar pattern: models that produce correct individual tool calls
frequently fail to maintain coherent task state across multi-turn
interactions~\cite{yao2023react}. The $\tau$-bench size sweep confirms that
this gap is not closed within the 0.5B--7B range: pass rates span 1.74\% to
4.35\% across all sizes and both configurations tested, with no consistent
scale dependence. The failure mode in the agentic setting is not format
compliance, which constrained decoding ensures, but multi-turn planning:
tracking which database fields have been retrieved, interpreting tool return
values, and selecting the correct next action across five to fifteen steps.
These capabilities scale with model size more slowly than single-call argument
extraction and are not addressed by any of the post-training or
inference-time techniques evaluated in this thesis.
